---
title: "`r tools::toTitleCase(params$target_type)` Signal Intensities within Groups"
date: "`r Sys.Date()`"
author : "The CompMS team"
editor_options: 
  chunk_output_type: console
params:
  pap: NULL
  factors: FALSE
  project_info: NULL
  target_type: "protein"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
---



```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 6.5,
  fig.align = "center",
  fig.pos = "H"
)

pap <- params$pap
lfqdataProt <-  pap$lfqdata_prot
tableconfig <- pap$lfqdata_prot$config$table
factors <- params$factors
project_info <- params$project_info

string_translations <- list(
  "metabolite" = list(
    "target" = "metabolite",
    "targets" = "metabolites",
    "proteins and peptides" = "metabolites",
    "feature_id" = "metabolite_feature_Id",
    "target:description" = "metabolite description"
  ),
  "peptide" = list(
    "target" = "peptide",
    "targets" = "peptides",
    "proteins and peptides" = "proteins and peptides"
  ),
  "protein" = list(
    "target" = "protein",
    "targets" = "proteins",
    "proteins and peptides" = "proteins",
    "feature_id" = "protein_Id",
    "target:description" = "protein description from fasta header"
  )
)
strings <- string_translations[[params$target_type]]

```


- Order : `r project_info$order_Id`
- Workunit : `r project_info$workunit_Id`

(ref:hierachycountspersample) Number of `r strings[["targets"]]` per Sample.

```{r hierachycountspersample, fig.cap="(ref:hierachycountspersample)"}
srs <- lfqdataProt$get_Summariser()
srs$plot_hierarchy_counts_sample()
```

```{r createPlot}
protID = tableconfig$hierarchy_keys_depth()

precabund <- pap$get_protein_per_group_abundance()
if (!factors) {
  precabund <- dplyr::filter(precabund , !!sym(tableconfig$factor_keys()[1]) == "ALL")
}



pp <- prolfquapp::plot_abundance_vs_percent(
  precabund,
  cfg_table = tableconfig,
  top_N = NULL,
  factors = factors,
  colors = c("^zz" = "red","Y-FGCZ" = "red"),
  columnAb = "abundance_percent",
  alpha = 0.4, logY = TRUE)
pp <- plotly::ggplotly(pp)
```


```{r createDT}
precabund_table <- as.data.frame(pap$get_protein_per_group_small_wide())
if (params$target_type == "metabolite") {
  # drop nrPeptides for metabolite
  precabund_table <- dplyr::select(precabund_table, -nrPeptides)
}
datax <- crosstalk::SharedData$new(precabund_table , key = as.formula(paste(" ~ ", protID )), group = "BB")

table <- DT::datatable(
  datax,
  filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE))


```

The columns of the table below contains:

- `r strings[["feature_id"]]` - `r strings[["target"]]` identifier
`r cat(ifelse(params$target_type == "metabolite", "", "- nrPeptides \n- nr of peptides per protein"))`
- description - `r strings[["target:description"]]`
- nrMeasured_<GroupName> - How often a `r strings[["target"]]` was observed in each of the groups and overall
- meanAbundance_<GroupName> - average `r strings[["target"]]` abundance (see 1. at the bottom of the document) per group.

-> TODO does this one actually still exist?
- signal_percent_<GroupName> - percent of the total signal of all identified `r strings[["targets"]]` measured by the mass spectrometric instrument attributed to each `r strings[["target"]]`.





```{r}
table
```


(ref:proteinCumulative) The Y axis presents the percentage of the iBAQ (intensity Based Absolute Quantitation) of each `r strings[["target"]]`. Along the X axis, `r strings[["targets"]]` are organized in ascending order according to their signal proportions. 



Calculating the iBAQ signal per `r strings[["target"]]` involves summing the `Precursor.Quantity` values across all precursors associated with that protein and dividing by the number of theoretically observable peptides of the protein. An intriguing observation is that a significant share of the mass spectrometers recorded signal (Y-axis) is allocated to a limited set of highly abundant proteins. Notably, when the most abundant proteins correspond to entities like the cleavage enzyme, typical contaminants such as human keratins, or the bait protein, this situation could indicate shortcomings in the sample processing.


```{r proteinCumulative, fig.cap = "(ref:proteinCumulative)"}
pp
```



```{r prepPeptidePlot, eval = FALSE, include=FALSE}
precabund <- pap$get_protein_per_group_abundance_with_row_annot()
datax <- crosstalk::SharedData$new(as.data.frame(precabund) , key = as.formula(paste(" ~ ", protID )), group = "BB")

colors = c("^zz" = "red", "Y-FGCZ" = "red")
colorV <- rep("black", nrow(precabund))
for (i in seq_along(colors)) {
  colorV[grepl(names(colors)[i], precabund[[protID]])] <- colors[i]
}
group <- tableconfig$factor_keys_depth()
pplotPepAbund <- ggplot(datax,
                        aes(x = nrPeptides, y = meanAbundance,
                            label = !!rlang::sym(protID))) + geom_point(color = colorV, alpha = 0.4) +
  ggplot2::scale_y_continuous(trans = 'log10') +
  ggplot2::scale_x_continuous(trans = 'log10') +
  if (factors) {facet_wrap(as.formula(paste0("~", paste(group, collapse = " + "))))}

pplotPepAbund <- plotly::ggplotly(pplotPepAbund)
```

(ref:nrProtplots) Mean `r strings[["target"]]` abundance per group (column meanAbundance_<GroupName>) as a function of the number of peptides (column nrPeptides).

```{r nrProtplots, fig.cap = "(ref:nrProtplots)", eval = FALSE, include=FALSE}
pplotPepAbund
```




