---
title: "Differential Expression Analysis."
author: "Functional Genomics Center Zurich"
date: today
format:
  html:
    toc-location: left
    grid:
      sidebar-width: 300px
    include-after-body: fgcz-plot-finder.html
toc: true
params:
  deanalyse_file: null
vignette: >
  %\VignetteIndexEntry{Differential Expression Analysis}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
deanalyse_file <- params$deanalyse_file
if (is.null(deanalyse_file) || !nzchar(deanalyse_file)) {
  deanalyse_file <- file.path(tempdir(), "prolfquapp-example-deanalyse.rds")
  if (!file.exists(deanalyse_file)) {
    saveRDS(prolfquapp::example_deanalyse(), deanalyse_file)
  }
}
dea <- readRDS(deanalyse_file)
cfg <- dea$prolfq_app_config
is_saint_report <- identical(dea$default_model, "saint")
empty_report_plot <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = label) +
    ggplot2::theme_void()
}
plot_or_empty <- function(plot, label) {
  if (is.null(plot)) {
    return(empty_report_plot(label))
  }
  plot
}

```


# B-fabric related information

This report is stored in the LIMS system [_bfabric_](https://fgcz-bfabric.uzh.ch) [@panse2022bridging] in
__project__: `r cfg$project_spec$project_Id`, __order__: `r cfg$project_spec$order_Id`, `r if(!is.null(cfg$project_spec$workunit_Id)){ "with the workunit name:" }`  `r cfg$project_spec$workunit_Id`.


The protein identification and quantification were performed using:
`r if(!is.null(cfg$software)){cfg$software} else {"<Specify parameter cfg$software>"}`. The input file can be downloaded from here: [`r cfg$project_spec$input_URL`](`r cfg$project_spec$input_URL`). The set of parameters used to run the quantification software can be retrieved from the b-fabric workunit to which this file belongs.

```{r introduction}
#| results: asis
#| eval: !expr "!is_saint_report"
cat(r"(
# Introduction

The differential expression analysis verifies if the __difference__ between normalized empirical protein abundances measured in two groups is significantly non-zero.
To make the test as sensitive and specific as possible, the methods used to measure [@taverna2021critical] and estimate protein abundances [@grossmann2010implementation] are optimized to minimize the biochemical and technical variance. In addition, these empirical abundances $A$ are further $log_2$ transformed and scaled to make them compatible with the statistical test procedure [@valikangas2018systematic]. Therefore, we obtain a scale free $\log_2$ transformed normalized protein abundances $\log_2(A)$ for a sample.

For unpaired experiments the difference $\Delta$ between group $a$ and $b$, for a specific protein is estimated by:

$$
\Delta = \frac{1}{n}\sum_{i=1}^n \log_2(A^a_i) - \frac{1}{m}\sum_{j=1}^{m}\log_2(A^b_j)
$$


where $A^a_i$ - is the normalized protein abundance of sample $i$ in the group $a$ of $n$ samples, while $A^b_j$ is the protein abundance of sample $j$ in group $b$ of $m$ samples.

For paired experiments, the difference is estimated by:


$$
\Delta = \frac{1}{n}\sum_{i=1}^n \log_2(A^a_i) - \log_2(A^b_{i})
$$


where $n$ is the number of subjects, each treated with $a$ and $b$.

Of note, when comparing two samples $a$ and $b$ the difference of logarithms equals the logarithm of the ratio (exponent rule):

$$
\log(A^a) - \log(A^b) = \log(\frac{A^a}{A^b})
$$


It is called the $\log$-ratio or $\log$ fold-change (logFC).

The estimated differences $\Delta$ have an associated error $\epsilon$. Therefore, the differential expression analysis must test if the difference is significantly nonzero.

We run a set of functions implemented in the R package _[prolfqua]_ [@prolfquawolski2023] to filter and normalize the data, generate visualizations, and to compute differential expression analysis. To further improve the power of the differential expression test the protein variances are moderated [@Smyth2004], i.e. the individual protein variances are updated using a variance prior estimated from all the proteins in the experiment.

)")
```

# Results

@tbl-samples shows the number of samples assigned to each group while @tbl-annotation shows the names of the files assigned to the group.

```{r}
#| label: tbl-samples
#| tbl-cap: "Nr of samples assigned to each group."
#| eval: true

tab <- data.frame(table(dea$lfq_data_raw$factors()[dea$lfq_data_raw$relevant_factor_keys()]))
colnames(tab)[length(colnames(tab))] <- "# samples"
knitr::kable(tab)

```


```{r}
#| label: tbl-annotation
#| tbl-cap: !expr paste0("LC-MS samples annotation table. The content of the sampleName column ", "is used as a short form to plot labels. The group to which a sample ", "is assigned is shown in the column group.")
#| eval: true
ld <- dea$lfq_data_raw
fx <- ld$factors()
sr <- ld$get_Summariser()
xx <- sr$hierarchy_counts_sample(nr_children = 1)
xx$isotopeLabel <- NULL
xx <- xx |> dplyr::rename(nr_1 = ld$hierarchy_keys())
x2 <- sr$hierarchy_counts_sample(nr_children = 2)
x2$isotopeLabel <- NULL
x2 <- x2 |> dplyr::rename(nr_2 = ld$hierarchy_keys())

fx <- dplyr::left_join(fx,xx, by = ld$sample_name())
fx <- dplyr::left_join(fx,x2, by = ld$sample_name())


knitr::kable(fx)

```

## Peptide and Protein identification

The protein matrix is filtered using the following threshold:

- Minimum number of peptides / protein: `r cfg$processing_options$nr_peptides`.

The overall number of proteins in this experiment is: `r dea$summary$totalNrOfProteins`.
The percentage of contaminant proteins is:`r dea$summary$percentOfContaminants` %.
The percentage of decoy proteins among quantified entries is: `r dea$summary$percentOfDecoys` %.

We keep the contaminant proteins because, for some experiments, these contaminants are relevant. They are recognizable by identifiers starting with __zz__ or __CON__, are flagged in the result tables, and are drawn as triangles in the volcano plot.
Decoy sequences (identifiers starting with __REV__ or __rev__) are retained in the raw and abundance exports for traceability, but they are **excluded from the model fit** so they cannot distort the variance moderation that is shared across proteins. They therefore carry no fold change, p-value, or FDR, and do not appear in the volcano plot or the differentially-expressed set. Their proportion among quantified entries (reported above) serves as an empirical false-discovery signal for the whole quantified data set — not for the differentially-expressed list.


@fig-nr-per-sample shows the number of quantified proteins with one or more peptides per sample.

```{r}
#| label: fig-nr-per-sample
#| fig-cap: "Number of quantified proteins per sample, counting proteins identified with at least one peptide; bar height is the protein count per sample."
#| fig-width: 10
#| fig-height: 7
sum <- dea$lfq_data_raw$get_Summariser()
sum$plot_hierarchy_counts_sample(nr_children = 1)

```

```{r prepNrPerSample2}
#| include: false
sum <- dea$lfq_data_raw$get_Summariser()
p_nr_per_sample_2 <- tryCatch(
  sum$plot_hierarchy_counts_sample(nr_children = 2),
  error = function(e) NULL
)
```

@fig-nr-per-sample-2 shows the number of quantified proteins with two or more peptides per sample.

```{r}
#| label: fig-nr-per-sample-2
#| fig-cap: "Number of quantified proteins per sample, counting only proteins identified with at least two peptides; bar height is the protein count per sample."
#| fig-width: 10
#| fig-height: !expr if (is.null(p_nr_per_sample_2)) 1.2 else 7
plot_or_empty(
  p_nr_per_sample_2,
  "No features with two or more children are available for this data set."
)

```


## Missing Value Analysis

The absence of a protein measurement in a sample might be biologically relevant or might point to technical problems. Significant differences in the set of proteins observed in the samples within a group typically indicate either technical problems or excessive biological variability. If one sample out of ten has a different set of proteins, it is likely an outlier and can be removed from the analysis. If the differences between the groups are significant but within the groups are small, this might systematically bias the difference estimates, i.e., produce false-positive or false-negative test results.


```{r countProtWithNAs}
res <- dea$lfq_data_raw$data_wide(as.matrix = TRUE)$data
res[!is.na(res)] <- 0
res[is.na(res)] <- 1
allrows <- nrow(res)
res <- res[apply(res,1, sum) > 0, , drop = FALSE]
allNas <- nrow(res)
```


```{r prepHeatmap}
pl <- dea$lfq_data_raw$get_Plotter()
nah <- tryCatch(
  pl$na_heatmap(),
  error = function(e) NULL
)


```

A dichotomous view of the data can be constructed by transforming protein abundance estimates into present/absent calls (@fig-na-heat). The heatmap shows only proteins with at least one missing value. There are `r allNas` proteins with at least one missing value in the data, which is (`r allNas/allrows * 100` %).

We expect that samples in the same group are more similar and cluster together, i.e., they are in the same branch of the dendrogram.

```{r}
#| label: fig-na-heat
#| fig-cap: "Protein abundance heatmap (rows indicate proteins, columns indicate samples) showing missing protein abundance estimates across the data set. Rows and columns are grouped based on the Minkowski distance using hierarchical clustering. White: Protein is observed, black: Protein is not observed."
#| fig-width: 7
#| fig-height: !expr if (is.null(nah)) 1.2 else 6
#| fig-dpi: 300
#| fig-alt: ""
#| eval: true
plot_or_empty(
  nah,
  "No missing values are available for the missing-value heatmap."
)
```

Using @fig-venn-proteins we examine if we see the same proteins in each group. We say a protein is unobserved in the group if it is absent in all samples and is present otherwise. A significant overlap among groups allows more precise estimation of the protein abundance differences between the groups.

```{r}
#| label: fig-venn-proteins
#| fig-cap: "UpSet plot of protein presence across groups. Bar heights give the number of proteins observed in each group and in each group intersection; a protein counts as present in a group when observed in at least one of its samples."
#| fig-alt: ""
pups <- prolfqua::upset_interaction_missing_stats(dea$lfq_data_raw, tr = 1)
UpSetR::upset(pups$data, order.by = "freq", nsets = pups$nsets)

```


## Protein Abundance Analysis

The density plot (@fig-normalized left panel)  displays the protein abundance distribution for all data set samples. Major differences between samples could indicate that the individual protein abundance values are affected by technical biases. These biases might need to be corrected to separate them from biological effects. The right panel of @fig-normalized shows the distribution of the transformed and scaled normalized empirical protein abundances. Normalization is applied to remove systematic differences in protein abundances due to different sample concentrations or amounts of sample loaded on a column. However, in the presence of a large proportion of missing data, normalization potentially amplifies systematic errors.


```{r normalization}
#| results: asis
if (cfg$processing_options$transform == "robscale") {
  cat(
    "To do this the z-score of the $\\log_2$ transformed protein",
    "abundances are computed.",
    "Because we need to estimate the protein differences on the",
    "original scale, we have to multiply the $z$-score by the average",
    "standard deviation of all the $N$ samples in the experiment.",
    "After normalization all samples have an equal mean and variance",
    "and a similar distribution.")
} else if (cfg$processing_options$transform == "vsn") {
  cat("To do this the variance stabilizing normalization (vsn) was applied [@HuberVSN2002].")
} else if (cfg$processing_options$transform == "none") {
  cat(
    "However, in some circumstances it is advisable not to normalize",
    "the data, e.g. in case of affinity purification experiments,",
    "or when the requirements of sufficient similarity among samples",
    "are not met.")
}

```

```{r}
#| label: fig-normalized
#| fig-dpi: 96
#| fig-cap: "Kernel density function showing the distribution of protein abundances in all samples. Left panel: Empirical protein abundance of all samples in the dataset. Right panel: Normalized empirical protein abundances of all samples in the dataset."
#| fig-width: 8
#| fig-height: 5
#| fig-alt: ""
#| eval: true

pl <- dea$lfq_data_raw$get_Plotter()
showlegend <- nrow(dea$lfq_data_raw$factors()) <= 12
p1 <- pl$intensity_distribution_density(legend = showlegend)

plTransformed <- dea$lfq_data$get_Plotter()
p2 <- plTransformed$intensity_distribution_density(legend = showlegend)

prolfquapp::plotly_ggplot_subplot(p1, p2, showlegend = showlegend, height = 450)

```

```{r internalCalibration}
#| results: asis
internal_ids <- cfg$processing_options$internal
if (length(internal_ids) > 0 && !all(is.na(internal_ids)) && !all(internal_ids == "")) {
  cat("In addition to the global normalization described above, **internal standard normalization** was applied.")
  cat(" The following protein(s) were used as internal reference standard(s):\n\n")
  cat(paste0("- `", internal_ids, "`", collapse = "\n"), "\n\n")
  cat("After transformation, the per-sample intensity of each reference",
    "protein was subtracted from all proteins in that sample",
    "(`center_to_reference`). ")
  cat("As a result, the reference protein(s) have an abundance of exactly 0 in the normalized data, ")
  cat("and all other protein abundances are expressed relative to the reference.\n\n")
}
```

The median coefficient of variation (CV) of a group of samples or all samples in the dataset, can be used to compare the experiment with other experiments [@piehowski2013sources]. For example, the median CV for high-performance liquid chromatography experiments ranges from 2% to 35% depending on the biological samples studied, the chromatography method used, label-free or labelled quantification [@taverna2021critical].


@fig-sd-violin shows the coefficients of variation (CV) for all proteins computed on non-normalized data. Ideally the within group CV should be smaller than the CV of all samples.

```{r}
#| label: fig-sd-violin
#| fig-cap: "Coefficient of variation (CV) of non-normalized protein abundances, shown as a violin per group and across all samples (all); the overlaid point marks the median CV."
#| fig-height: 3
#| fig-width: 7
#| fig-alt: ""
#| eval: true
stR <- dea$lfq_data_raw$get_Stats()
pA <- stR$violin() + ggplot2::theme_classic() + ggplot2::labs(tag = "A")
pA

```

@tbl-cv shows the median CV of all groups and across all samples (all).

```{r}
#| label: tbl-cv
#| tbl-cap: "Median coefficient of variation (CV) of non-normalized protein abundances, per group and across all samples (all)."
#| eval: true
resR <- stR$stats_quantiles()
C <- dplyr::bind_rows(CV = resR$wide  |> dplyr::filter(probs == 0.5) |> round(digits = 2))
C <- C |> dplyr::mutate( what = c("CV"), .before = 1 )
C$probs <- NULL
knitr::kable(C)

```



```{r generateHeatmaps}
#| include: false
pl <- dea$lfq_data$get_Plotter()
ph <- pl$heatmap()

```

The protein abundance heatmap (@fig-heatmap) groups the protein and samples using unsupervised hierarchical clustering. Distances between proteins and samples are computed using normalized protein abundances. Proteins with a large proportion of missing observations are not shown in this heatmap, because for these proteins no distance can be computed.  Proteins and samples showing similar abundances are grouped and shown in adjacent rows and columns respectively.


```{r}
#| label: fig-heatmap
#| fig-cap: "Protein abundance heatmap (rows indicate proteins, columns indicate samples) showing the row scaled $\\log_2$ transformed protein abundance value. Co-clustering (hierarchical complete linkage, euclidean distance) of samples and proteins was used. Cell color encodes the row z-score on a green-black-red scale (green: low, black: average, red: high); missing values are shown in light gray."
#| fig-width: 7
#| fig-height: 8
#| fig-dpi: 300
#| fig-alt: ""
#| eval: true
ph
```

We use principal component analysis (PCA) to transform the high-dimensional space defined by all proteins into a two-dimensional one containing most of the information. @fig-pca shows the location of the samples according to the first and second principal component, which explain most of the variance in the data. Samples close in the PCA plot are more similar than those farther apart.

```{r}
#| label: fig-pca
#| fig-dpi: 96
#| fig-cap: "Plot of first and second principal component (PC1 and PC2) of principal component analysis (PCA). Normalized abundances were used as input."
#| fig-alt: ""
#| fig-width: 7
#| fig-height: 7
pca_widget <- dea$lfq_data$get_Plotter()$pca_plotly()
pca_widget$width <- "100%"
pca_widget$height <- 500
pca_widget
```

# Differential Expression Analysis

The method used to test for differential expression consists of several steps: First a model that explains the observed protein abundances using the grouping of the samples is fitted to each protein with the selected backend (the `modelName`/facade key; by default a linear model fitted with the R function _lm_):

`r dea$formula`.

Secondly, the difference between the groups is computed (@tbl-contrasts).

```{r}
#| label: tbl-contrasts
#| tbl-cap: "Name of difference (Contrast), and formula used to compute it."
df <- data.frame(name = names(dea$contrasts), contrast = dea$contrasts)
rownames(df) <- NULL
knitr::kable(df)

```


and a null hypothesis significance test (NHST) is conducted, where the null hypothesis is that the protein is not differentially expressed [@faraway2004linear].

If there are no abundances measured in one of the groups for some proteins, we assume the observations are missing because the protein abundance is below the detection limit. Therefore, we estimate the detection limit using the mean of the $1\%$ smallest group averages.
Furthermore, to make it explicit how each estimate was produced, the `estimate_type` column (see table in @fig-table-all) records whether a row was fit from `observed` data, rescued via LOD imputation (`lod_imputed`), or filled from a group-mean fallback (`missing_fallback`); rescued/fallback rows are visualized with a distinct colour in @fig-volcano. The `modelName` column records the selected analysis method (the facade key, e.g. `lm`, `rfit`, `limma`).

Next, to increase the power of the analysis variance shrinkage is performed [@Smyth2004]. Finally, the false discovery rate (FDR) using the Benjamini-Hochberg procedure is computed [@benjamini1995controlling].
The FDR is the expected proportion of false discoveries in a list of proteins, and can be used to select candidates for follow up experiments. FDR thresholds commonly used are 5, 10 or 25%. By filtering the proteins using an FDR threshold of `r cfg$processing_options$FDR_threshold*100` % we can expect this proportion of false positives in the list and `r (1 - cfg$processing_options$FDR_threshold) * 100` % truly differentially expressed proteins. Because we do not know which of them are true positives follow up experiments are necessary.

The table (@fig-table-all) summarizes the differential expression analysis results by providing the following information:

```{r}
hkey <- dea$lfq_data$hierarchy_keys()
```

- `r if(hkey[1] == "protein_Id"){ paste0(hkey[1], " - unique protein identifier")} else {hkey[1]}`
- description - information about the protein provided in the FASTA database
- contrast - name of the comparison
- modelName - selected analysis method (facade key, e.g. `lm`, `rfit`, `limma`)
- estimate_type - how the estimate was produced: `observed`, `lod_imputed`, or `missing_fallback`
- FDR - false discovery rate
- diff - difference between groups.

The volcano plot @fig-volcano helps to identify proteins with large differences among groups and a low FDR. The significance dimension is a $-\log_{10}$ transformed FDR, i.e., small values of FDR become large after transformation. Promising candidate proteins are found in the upper right and left sector of the plot.


```{r setupVolcano, echo = FALSE}
datax <- as.data.frame(dea$annotated_contrasts)
is_saint <- identical(dea$default_model, "saint")
contrast_col <- if (is_saint) "Bait" else "contrast"
effect_col <- if (is_saint) "log2_EFCs" else "diff"
score_col <- if (is_saint) "BFDR" else "FDR"
table_cols <- if (is_saint) {
  c("ID", "description", "Bait", "modelName", "log2_EFCs", "BFDR")
} else {
  c("ID", "description", "contrast", "modelName", "estimate_type", "diff", "FDR")
}
volcano_cap <- paste0(
  "Volcano plot showing $-\\log_{10}$ transformed FDR as function of the difference ",
  "between groups. The red line indicates the $-log_{10}(FDR)$ of FDR = ",
  cfg$processing_options$FDR_threshold,
  ", while the green lines represent the difference of minus and plus ",
  cfg$processing_options$diff_threshold,
  ". Points are coloured by `estimate_type`, so rows estimated by missing-value ",
  "imputation or group-mean fallback are visually distinguished from observed fits."
)
```


```{r}
#| label: fig-table-all
#| fig-cap: "Interactive, filterable table of per-protein, per-contrast differential expression results: protein ID, description, contrast, model, estimate type, log2 fold-change and FDR."

datax <- datax |> tidyr::unite("ID" , tidyselect::all_of(hkey), sep = "~")

bb <- datax |> dplyr::select(
  tidyselect::all_of(table_cols))

bb <- crosstalk::SharedData$new(bb,  as.formula("~ ID"), group = "BB")

DT::datatable(bb, filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE) ) |>
  DT::formatRound(columns = c(score_col, effect_col), digits = 3)

```

```{r}
#| label: fig-volcano
#| fig-dpi: 96
#| fig-cap: !expr volcano_cap
#| fig-alt: ""
#| fig-width: 9
#| fig-height: 7
#| eval: true

plotter <- if (is_saint) {
  dea$contrast_results[[dea$default_model]]$get_Plotter(
    fc_threshold = cfg$processing_options$diff_threshold,
    bfdr_threshold = cfg$processing_options$FDR_threshold
  )
} else {
  dea$contrast_results[[dea$default_model]]$get_Plotter(
    fc_threshold = cfg$processing_options$diff_threshold,
    fdr_threshold = cfg$processing_options$FDR_threshold
  )
}
volcano_widget <- plotter$volcano_plotly()$FDR
volcano_widget$width <- "100%"
volcano_widget$height <- 500
volcano_widget

```

## Differentially Expressed Proteins

Here we use the FDR threshold of `r cfg$processing_options$FDR_threshold` and a difference threshold of `r cfg$processing_options$diff_threshold` to select differentially expressed proteins. @tbl-nr-significant summarizes the number of significant calls.

```{r}
#| label: tbl-nr-significant
#| tbl-cap: "Number of proteins called significant versus not significant per contrast, applying the fold-change and FDR thresholds (total, significant, not significant)."
#| eval: true
library(dplyr)
mx <- datax |>
  dplyr::mutate(
    passes = if (is_saint) {
      log2_EFCs > cfg$processing_options$diff_threshold &
        BFDR < cfg$processing_options$FDR_threshold
    } else {
      abs(diff) > cfg$processing_options$diff_threshold &
        FDR < cfg$processing_options$FDR_threshold
    })
x <- mx |>
  dplyr::group_by(.data[[contrast_col]]) |>
  summarize(
    n = n(),
    Significant = sum(passes, na.rm = TRUE),
    "Not Significant" = n() - sum(passes, na.rm = TRUE))

knitr::kable(x)

```



```{r areThereSig}
datax_signif <- dea$annotated_contrasts_signif
showSignificant <- TRUE
if (nrow(datax_signif) == 0) {
  showSignificant <- FALSE
}

```


```{r}
#| results: asis
#| eval: !expr showSignificant
cat("The table shown in @fig-sig-prey lists all the significant proteins.")
```

```{r}
#| label: fig-sig-prey
#| fig-cap: "Significant proteins obtained by applying the difference and FDR thresholds."
#| fig-alt: ""
#| eval: !expr showSignificant
ctdata <- datax_signif |> tidyr::unite("ID", tidyselect::all_of(hkey), sep = "~")
ctdata <- ctdata |>
  dplyr::select(tidyselect::all_of(table_cols)) |>
  as.data.frame()

sig <- crosstalk::SharedData$new(ctdata,  as.formula( "~ ID"), group = "BB")
DT::datatable(sig, filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE) ) |>
  DT::formatRound(columns = c(score_col, effect_col), digits = 3)
```




```{r prepareHeatmap}
#| eval: !expr showSignificant
signif <- dea$lfq_data$get_copy()
signif <- signif$get_subset(datax_signif)
if (signif$hierarchy_counts()[2] > 30) {
  rownames = FALSE
} else {
  rownames = TRUE
}
sigheat <- signif$get_Plotter()$heatmap(rownames = rownames)

```


```{r makeText}
#| eval: !expr showSignificant
#| results: asis

cat(paste(
  "Furthermore, @fig-sig-proteins shows a co-clustered heatmap",
  "of the row z-scored protein abundances of all significant calls.",
  "Proteins and samples with similar abundance profiles are grouped by",
  "unsupervised hierarchical clustering."))

```

```{r}
#| label: fig-sig-proteins
#| fig-cap: "Heatmap of the proteins passing the FDR and difference thresholds. Cell colour encodes the row z-score of the $\\log_2$ protein abundance (green: low, black: average, red: high; missing values in light gray). Proteins and samples are co-clustered by unsupervised hierarchical clustering (complete linkage, euclidean distance)."
#| fig-alt: ""
#| eval: !expr showSignificant
sigheat
```

```{r}
#| label: fig-venn-sig
#| fig-cap: "UpSet plot of significant proteins across contrasts. Bar heights give the number of proteins passing the FDR and fold-change thresholds in each contrast and in the contrast intersections."
#| eval: !expr showSignificant
xx <- split(ctdata[["ID"]], ctdata[[contrast_col]])
if (length(xx) > 1) {
  UpSetR::upset(UpSetR::fromList(xx))
}

```


# Additional Analysis

The zip file contains an Excel file __DE_Groups_vs_Controls.xlsx__. All the figures can be recreated using the data in the excel file. The Excel file contains the following spreadsheets:

- __annotation__ - the annotation of the samples in the experiment
- __raw_abundances__ table with empirical protein abundances.
- __normalized_abundances__ table with normalized protein abundances.
- __raw_abundances_matrix__ A table where each column represents a sample and each row represents a protein and the cells store the empirical protein abundances.
- __normalized_abundances_matrix__ A table where each column represents a sample and each row represents a protein and the cells store the empirical protein abundances.
- __diff_exp_analysis__ A table with the results of the differential expression analysis. For each protein there is a row containing the estimated difference between the groups, the false discovery rate FDR, the 95% confidence interval, the posterior degrees of freedom.
- __missing_information__ - spreadsheet containing information if a protein is present (1) or absent in a group (0).
- __protein_variances__ - spreadsheet which for each protein shows the variance (var) or standard deviation (sd) within a group, the number of samples (n) and the number of observations (not_na) as well as the group average intensity (mean).


The data can be used to perform functional enrichment analysis [@monti2019proteomics].
To compare the obtained results with known protein interactions we recommend the [string-db.org](https://string-db.org/) [@Szklarczyk2017], which is a curated database of protein-protein interaction networks for a large variety of organisms. To simplify the data upload to string-db we include text files containing the uniprot ids:


- `ORA_background.txt` all proteins.
- `ORA_<contrast_name>.txt` proteins accepted with the FDR and difference threshold.

Other web applications allowing to run over representation analysis (ORA) [@monti2019proteomics] are:

- [DAVID Bioinformatics Resource](https://david.ncifcrf.gov/home.jsp)
- [WEB-based GEne SeT AnaLysis Toolkit](http://www.webgestalt.org) [@Wang2017]


Furthermore, protein IDs sorted by t-statistic can then be subjected to gene set enrichment analysis (GSEA) [@subramanian2005gene].
To simplify running GSEA, we provide the file:

- `GSEA_<contrast_name>.rnk`

This file can be used with the webgestalt web application or used with the GSEA application from [gsea-msigdb](https://www.gsea-msigdb.org/)


For questions and improvement suggestions, with respect to this report, please contact <protinf@fgcz.uzh.ch>.


# Session Information

```{r sessionInfo}
pander::pander(sessionInfo())
```


# References {.unnumbered}

::: {#refs}
:::


# Glossary

- groups - different treatments, genotypes etc.
- diff (difference) - it is the difference of the protein abundance estimate of two groups
- FDR - false discovery rate


---

*This report was generated from the Quarto template `Grp2Analysis_V2_R6.qmd` included in the `prolfquapp` R package (version `r packageVersion("prolfquapp")`).*
