---
title: "Protein Signal Intensities within Groups"
author: "The CompMS team"
date: today
format:
  html:
    grid:
      sidebar-width: 300px
    include-after-body: fgcz-plot-finder.html
toc: false
params:
  pap_file: null
  factors: true
  project_info: null
vignette: >
  %\VignetteIndexEntry{QC Protein Abundances}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 6.5,
  fig.align = "center"
)

pap_file <- params$pap_file
if (is.null(pap_file) || !nzchar(pap_file)) {
  pap_file <- file.path(tempdir(), "prolfquapp-example-qc-generator.rds")
  if (!file.exists(pap_file)) {
    saveRDS(prolfquapp::example_qc_generator(), pap_file)
  }
}
pap <- readRDS(pap_file)
lfqdataProt <- pap$lfqdata_prot
tableconfig <- pap$lfqdata_prot$get_config()
factors <- params$factors
project_info <- params$project_info

```


- Order : `r project_info$order_Id`
- Workunit : `r project_info$workunit_Id`

::: {.panel-tabset}

# Protein abundances

```{r createPlot}
protID <- tableconfig$hierarchy_keys_depth()

precabund <- pap$get_protein_per_group_abundance()
if (!factors) {
  precabund <- dplyr::filter(precabund, !!sym(tableconfig$factor_keys()[1]) == "ALL")
}



pp <- prolfquapp::plot_abundance_vs_percent(
  precabund,
  cfg_config = tableconfig,
  top_N = NULL,
  factors = factors,
  colors = c("^zz" = "red", "Y-FGCZ" = "red"),
  columnAb = "abundance_percent",
  alpha = 0.4,
  highlight_alpha = 0.8,
  logY = TRUE)
pp <- plotly::ggplotly(pp)
```


```{r createDT}
precabund_table <- pap$get_protein_per_group_small_wide()
datax <- crosstalk::SharedData$new(
  as.data.frame(precabund_table),
  key = as.formula(paste(" ~ ", protID)),
  group = "BB")

table <- DT::datatable(
  datax,
  filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(
    deferRender = TRUE,
    scrollY = 300,
    scrollX = 400,
    scroller = TRUE
  ))


```

The columns of the table below contain:

- protein_Id - protein identifier
- nrPeptides - number of peptides assigned to the protein
- description - protein description from the FASTA header
- nrMeasured_<GroupName> - number of samples in which the protein was quantified, by group and overall
- meanAbundance_<GroupName> - mean protein abundance per group
- signal_percent_<GroupName> - percentage of the total protein signal attributed to the protein




```{r}
table
```


The iBAQ signal per protein is calculated by summing the `Precursor.Quantity` values across all precursors assigned to the protein and dividing by the number of theoretically observable peptides. A large share of the recorded signal can be concentrated in a small set of highly abundant proteins. If the dominant proteins are the cleavage enzyme, common contaminants such as human keratins, or the bait protein, this can indicate shortcomings in sample preparation or cleanup.


```{r}
#| label: fig-protein-cumulative
#| fig-dpi: 96
#| fig-cap: "Per-protein iBAQ signal contribution versus abundance rank, one panel per group. Each point is one protein; the x-axis is the protein's abundance-rank percentile (proteins ordered from lowest to highest mean iBAQ abundance) and the y-axis (log10 scale) is the percentage of the total iBAQ (intensity-based absolute quantification) signal contributed by that protein. Contaminant and decoy proteins are highlighted in red."
pp
```

# Proteins per sample

```{r}
#| label: fig-hierarchy-counts-sample
#| fig-cap: "Number of proteins quantified per sample."
srs <- lfqdataProt$get_Summariser()
srs$plot_hierarchy_counts_sample()
```

:::
