---
title: "Protein Signal Intensities within Groups"
date: "`r Sys.Date()`"
author: "The CompMS team"
editor_options:
  chunk_output_type: console
params:
  pap: !r quote(prolfquapp::example_qc_generator(Nprot = 100))
  factors: TRUE
  project_info: NULL
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_depth: 2
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="prolfquapp")
    css: !expr system.file("templates/fgcz.css", package="prolfquapp")
vignette: >
  %\VignetteIndexEntry{QC Protein Abundances}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---



```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 6.5,
  fig.align = "center",
  fig.pos = "H"
)

if (is.language(params$pap)) {
  pap <- eval(params$pap)
} else {
  pap <- params$pap
}
lfqdataProt <-  pap$lfqdata_prot
tableconfig <- pap$lfqdata_prot$get_config()
factors <- params$factors
project_info <- params$project_info

```


- Order : `r project_info$order_Id`
- Workunit : `r project_info$workunit_Id`

```{r hierachycountspersample, fig.cap="Number of proteins quantified per sample."}
srs <- lfqdataProt$get_Summariser()
srs$plot_hierarchy_counts_sample()
```

```{r createPlot}
protID <- tableconfig$hierarchy_keys_depth()

precabund <- pap$get_protein_per_group_abundance()
if (!factors) {
  precabund <- dplyr::filter(precabund, !!sym(tableconfig$factor_keys()[1]) == "ALL")
}



pp <- prolfquapp::plot_abundance_vs_percent(
  precabund,
  cfg_config = tableconfig,
  top_N = NULL,
  factors = factors,
  colors = c("^zz" = "red", "Y-FGCZ" = "red"),
  columnAb = "abundance_percent",
  alpha = 0.4,
  logY = TRUE)
pp <- plotly::ggplotly(pp)
```


```{r createDT}
precabund_table <- pap$get_protein_per_group_small_wide()
datax <- crosstalk::SharedData$new(
  as.data.frame(precabund_table),
  key = as.formula(paste(" ~ ", protID)),
  group = "BB")

table <- DT::datatable(
  datax,
  filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(
    deferRender = TRUE,
    scrollY = 300,
    scrollX = 400,
    scroller = TRUE
  ))


```

The columns of the table below contain:

- protein_Id - protein identifier
- nrPeptides - number of peptides assigned to the protein
- description - protein description from the FASTA header
- nrMeasured_<GroupName> - number of samples in which the protein was quantified, by group and overall
- meanAbundance_<GroupName> - mean protein abundance per group
- signal_percent_<GroupName> - percentage of the total protein signal attributed to the protein





```{r}
table
```


(ref:proteinCumulative) Protein-level signal contribution. Each point represents one protein. Proteins are ordered on the x-axis by their contribution to the total signal; the y-axis shows the percentage of the total iBAQ (intensity-based absolute quantification) signal attributed to each protein.


The iBAQ signal per protein is calculated by summing the `Precursor.Quantity` values across all precursors assigned to the protein and dividing by the number of theoretically observable peptides. A large share of the recorded signal can be concentrated in a small set of highly abundant proteins. If the dominant proteins are the cleavage enzyme, common contaminants such as human keratins, or the bait protein, this can indicate shortcomings in sample preparation or cleanup.


```{r proteinCumulative, fig.cap = "(ref:proteinCumulative)"}
pp
```



```{r prepPeptidePlot, eval = FALSE, include=FALSE}
precabund <- pap$get_protein_per_group_abundance_with_row_annot()
datax <- crosstalk::SharedData$new(as.data.frame(precabund) , key = as.formula(paste(" ~ ", protID )), group = "BB")

colors <- c("^zz" = "red", "Y-FGCZ" = "red")
colorV <- rep("black", nrow(precabund))
for (i in seq_along(colors)) {
  colorV[grepl(names(colors)[i], precabund[[protID]])] <- colors[i]
}
group <- tableconfig$factor_keys_depth()
pplotPepAbund <- ggplot(datax,
                        aes(x = nrPeptides, y = meanAbundance,
                            label = !!rlang::sym(protID))) + geom_point(color = colorV, alpha = 0.4) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::scale_x_continuous(trans = "log10") +
  if (factors) {facet_wrap(as.formula(paste0("~", paste(group, collapse = " + "))))}

pplotPepAbund <- plotly::ggplotly(pplotPepAbund)
```

(ref:nrProtplots) Mean protein abundance per group (column meanAbundance_<GroupName>) as a function of the number of peptides (column nrPeptides).

```{r nrProtplots, fig.cap = "(ref:nrProtplots)", eval = FALSE, include=FALSE}
pplotPepAbund
```

