---
title: "Quality Control & Sample Size Estimation"
author: "WEW@FGCZ.ETHZ.CH"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_depth: 2
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package="prolfquapp")
    css: !expr system.file("templates/fgcz.css", package="prolfquapp")
params:
  configuration:  !r quote(prolfqua::prolfqua_data('data_ionstar')$filtered()$config$clone(deep=TRUE))
  project_conf: NULL
  data: !r quote(prolfqua::prolfqua_data('data_ionstar')$filtered()$data)
  plot_density: TRUE
  plot_sd_vs_mean: FALSE
  target_type: "protein"
vignette: >
  %\VignetteIndexEntry{Quality Control & Sample Size Estimation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  editor_options:
    chunk_output_type: console
  chunk_output_type: console
header-includes:
  - \usepackage[english]{babel}
  - \usepackage{blindtext}
  - \usepackage{float}
---




```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5.5,
  fig.height = 3.5,
  fig.align = "center",
  fig.pos = "H"
)

data <- eval(params$data)

local_project_conf <- params$project_conf
local_config <- eval(params$configuration)

if (is.null(local_project_conf)) {
  local_project_conf <- list()
}

if (is.null(params$plot_density)) {
  params$plot_density <- TRUE
}

if (is.null(params$plot_sd_vs_mean)) {
  params$plot_sd_vs_mean <- FALSE
}
old.theme <- ggplot2::theme_set(ggplot2::theme_classic())


string_translations <- list(
  "metabolite" = list(
    "target" = "metabolite",
    "targets" = "metabolites",
    ".analyzed_target_types" = "metabolites"
  ),
  "peptide" = list(
    "target" = "peptide",
    "targets" = "peptides",
    ".analyzed_target_types" = "proteins and peptides"
  ),
  "protein" = list(
    "target" = "protein",
    "targets" = "proteins",
    ".analyzed_target_types" = "proteins"
  )
)
strings <- string_translations[[params$target_type]]

```

\clearpage

# Introduction

- Workunit: `r local_project_conf$workunit_Id`
- Project: `r local_project_conf$project_Id`
- Order : `r local_project_conf$order_Id`

<!--This document does not contain any specific conclusion since it was generated automatically. Hence, please discuss this document with your project bioinformatician/statistician which typically is not the project coach.-->

We did run your samples through the same analysis pipeline, which will be applied in the main experiment.
This document summarizes the `r strings[["target"]]` variability to assess the reproducibility of the biological samples and estimates the sample sizes needed for the main experiment.


# Quality Control: Identifications and Quantifications

```{r typicalObservations, echo=FALSE}
if (params$target_type == "protein" || params$target_type == "peptide") {
  typical_observations <- "Depending on the type of your sample (e.g., pull-down, supernatant, whole cell lysate) we observe some dozens up to a few thousands of proteins"
  if (params$target_type == "peptide") {
    typical_observations <- paste0(typical_observations, ", and between a few hundred up to some few tens of thousands of peptides.")
  } else {
    typical_observations <- paste0(typical_observations, ".")
  }
} else {
  # Not defined for metabolite
  typical_observations <- ""
}
```

Here we summarize the number of `r strings[["targets"]]` measured in the QC experiment.
`r typical_observations`
While the overall number of `r strings[[".analyzed_target_types"]]` can highly vary depending of the type of experiment, it is crucial that the number of `r strings[[".analyzed_target_types"]]`  between your biological replicates is similar (reproducibility).


```{r hierarchyCounts}
library(rlang)
library(prolfqua)
lfqdata <- LFQData$new(data, local_config)

x <- lfqdata$hierarchy_counts()
prolfqua::table_facade( 
  data.frame(NR = x),
  caption = paste0("Nr of ", strings[[".analyzed_target_types"]], " detected in all samples."))

```



```{r}

nrSamples <- nrow(lfqdata$factors())
width = max(8, nrSamples*0.2)
```

(ref:hierarchyCountsSampleBarplot) Number of quantified `r strings[["targets"]]` per sample. Each bar represents one LC-MS sample and the y-axis shows the count of quantified `r strings[["targets"]]`.


```{r hierarchyCountsSampleBarplot, fig.cap="(ref:hierarchyCountsSampleBarplot)", fig.width=width, fig.height=6}
summarizer <- lfqdata$get_Summariser()
summarizer$plot_hierarchy_counts_sample()
```

(ref:hierarchyCountsSample) Overlap of quantified `r strings[["targets"]]` across samples. Bars show set and intersection sizes, and filled dots mark which samples contribute to each intersection.

```{r upsetMissing, fig.width=9, fig.height=9, fig.cap="(ref:hierarchyCountsSample)"}
plotter <- lfqdata$get_Plotter()
plotter$upset_missing()

```



Ideally, we identify each `r strings[["target"]]` in all of the samples. However, because of the limit of detection (LOD) low-intensity `r strings[["targets"]]` might not be observed in all samples. Ideally, the LOD should be the only source of missingness in biological replicates. The following figures help us to verify the reproducibility of the measurement at the level of missing data.

```{r}
height <- max(7, nrSamples * 0.1)
```

(ref:missingFigIntensityHistorgram) Missing-value structure across `r strings[["target"]]` quantifications. The upper panel shows intensity distributions grouped by the number of missing values, and the lower panels show counts and cumulative counts of missing values per group.

```{r missingFigIntensityHistorgram, fig.width=height, fig.height=height, fig.cap="(ref:missingFigIntensityHistorgram)"}

p <- plotter$missigness_histogram()
xx3 <- summarizer$plot_missingness_per_group()
xx4 <- summarizer$plot_missingness_per_group_cumsum()
gridExtra::grid.arrange(p, xx3, xx4, ncol = 1)

```

(ref:missingnessHeatmap) Missing `r strings[["target"]]` quantifications clustered by sample. Rows are `r strings[["targets"]]`, columns are samples, and the heatmap marks observed and missing measurements.

```{r preparemissingnessHeatmap, include=FALSE}
pNAmap <- plotter$na_heatmap()
width <- max(8, nrSamples * 0.15)
```


```{r missingnessHeatmap, fig.width=width, fig.height=width, fig.align='center', fig.cap="(ref:missingnessHeatmap)"}
print(pNAmap)

```


```{r checktransformation, include = FALSE}
show_text <- !lfqdata$is_transformed()
```

```{r transformIntensities, include = FALSE}
if (show_text) {
  tr <- lfqdata$get_Transformer()
  n_samples <- length(unique(lfqdata$data_long()[[lfqdata$sample_name()]]))
  if (n_samples < 2) {
    logger::log_warn("vsn requires >= 2 samples, falling back to log2 transformation.")
    tr$log2()
  } else {
    tr$intensity_matrix(.func = vsn::justvsn)
  }
  dataTransformed <- tr$lfq
} else {
  dataTransformed <- lfqdata
}
plotter_transformed <- dataTransformed$get_Plotter()
```

```{r variability, eval = show_text, results='asis' }
cat("## Variability of raw intensities\n\n")

cat(
  "Before intensity scaling and preprocessing, the ", strings[["target"]],
  " intensities should have comparable distributions across samples. We assess ",
  "this with ", strings[["target"]], "-level coefficient of variation (CV) densities ",
  "(Figure \\@ref(fig:plotDistributions)) and sample-level raw versus transformed ",
  "intensity distributions (Figure \\@ref(fig:intensityDistribution)). ",
  "Table \\@ref(tab:printTable) summarizes CV quantiles."
)

```

(ref:plotDistributions) Distribution of `r strings[["target"]]`-level coefficients of variation (CV) before transformation. The left panel shows CV densities by group, and the right panel splits `r strings[["targets"]]` into below- and above-median abundance strata.

```{r plotDistributions, fig.cap="(ref:plotDistributions)", fig.height=6, fig.width=8 , eval = show_text}

moreThanOneSample <- nrow(lfqdata$factors()) > 1

if ( moreThanOneSample ) {
  stats <- lfqdata$get_Stats()
  if (params$plot_density) {
    p1 <- stats$density() +
      ggplot2::labs(title = "All quantified targets", x = "CV", y = "density") +
      ggplot2::labs(tag = "A") +
      ggplot2::xlim(0, 150)
    p2 <- stats$density_median() +
      ggplot2::labs(title = "Below and above median abundance", x = "CV", y = "density") +
      ggplot2::labs(tag = "B") +
      ggplot2::xlim(0, 150)
    prolfquapp::plotly_ggplot_subplot(p1, p2)
  } else {
    p1 <- stats$violin() + ggplot2::labs(tag = 'A')
    p2 <- stats$violin_median() + ggplot2::labs(tag = 'B')
    gridExtra::grid.arrange(p1,p2)
  }
}


```

```{r computeCVQuantiles, include=FALSE, eval = show_text}
cv_quantiles_res <- stats$stats_quantiles(probs = c(0.5, 0.6, 0.7, 0.8, 0.9))$wide
```


```{r printTable, results="asis", eval = show_text}
prolfqua::table_facade(cv_quantiles_res,
                         caption = "Summary of the coefficient of variation (CV) at the 50th, 60th, 70th, 80th and 90th percentile.")


```

## Variability of transformed intensities

We applied the `vsn::justvsn` normalization to the data, which should remove systematic differences among the samples and reduce the variance within the groups. Figure \@ref(fig:intensityDistribution) shows the sample-level intensity distributions before and after transformation. Because of this transformation, we cannot report CVs on the transformed scale and instead report standard deviations (SD). Figure \@ref(fig:sdviolinplots) shows the distribution of `r strings[["target"]]` SDs, Figure \@ref(fig:sdecdf) shows their empirical cumulative distribution function (ECDF), and Table \@ref(tab:printSDTable) summarizes SD quantiles. The heatmap in Figure \@ref(fig:correlationHeat) shows correlations among the QC samples.


(ref:intensityDistribution) Sample-level intensity distributions before and after transformation. The left panel shows raw `r strings[["target"]]` intensities, the right panel shows transformed intensities, and each curve or violin represents one sample.

```{r intensityDistribution, fig.cap="(ref:intensityDistribution)", fig.height = 5, fig.width=8, eval = show_text}
if (params$plot_density) {
  raw_intensity <- lfqdata$get_Plotter()$intensity_distribution_density() +
    ggplot2::labs(title = "Raw intensities") +
    ggplot2::theme(legend.text = ggplot2::element_text(size = 5))
  transformed_intensity <- plotter_transformed$intensity_distribution_density() +
    ggplot2::labs(title = "Transformed intensities") +
    ggplot2::theme(legend.text = ggplot2::element_text(size = 5))
  prolfquapp::plotly_ggplot_subplot(
    raw_intensity,
    transformed_intensity,
    showlegend = nrow(lfqdata$factors()) <= 12
  )
} else {
  raw_intensity <- lfqdata$get_Plotter()$intensity_distribution_violin() +
    ggplot2::labs(title = "Raw intensities")
  transformed_intensity <- plotter_transformed$intensity_distribution_violin() +
    ggplot2::labs(title = "Transformed intensities")
  gridExtra::grid.arrange(raw_intensity, transformed_intensity, ncol = 2)
}
```

(ref:correlationHeat) Sample correlation heatmap after transformation. Rows and columns are samples, and colours encode pairwise correlations of transformed `r strings[["target"]]` intensities.

```{r preparecorrelationHeat, include= FALSE}
chmap <- plotter_transformed$heatmap_cor()
width <- max(8, nrSamples * 0.15)
```

```{r correlationHeat, fig.height = 5, fig.cap="(ref:correlationHeat)", fig.width=width, fig.height=width}
print(chmap)
```


```{r pairsplotSmooth, fig.cap = "Pairwise scatter plots of transformed sample intensities. Each panel compares two samples, and smoothed trends show agreement between samples.", fig.height=12, fig.width=12}
invisible(plotter_transformed$pairs_smooth(max = 10))
```




(ref:sdviolinplots) Distribution of `r strings[["target"]]` standard deviations after transformation. The left panel shows SD densities by group, and the right panel splits `r strings[["targets"]]` into below- and above-median abundance strata.

```{r sdviolinplots,fig.cap="(ref:sdviolinplots)", fig.height=6, fig.width=8}

st <- dataTransformed$get_Stats()

if (params$plot_density) {
  p1 <- st$density() +
    ggplot2::labs(title = "All quantified targets", tag = "A", x = "SD", y = "density")
  p2 <- st$density_median() +
    ggplot2::labs(title = "Below and above median abundance", tag = "B", x = "SD", y = "density")
  prolfquapp::plotly_ggplot_subplot(p1, p2)
} else {
  p1 <- st$violin() +
    ggplot2::labs(tag = "A") +
    ggplot2::theme(legend.position = "none")
  p2 <- st$violin_median() +
    ggplot2::labs(tag = "B") +
    ggplot2::theme(legend.position = "bottom")
  gridExtra::grid.arrange(p1, p2)
}
```


(ref:sdecdf) Empirical cumulative distribution function (ECDF) of `r strings[["target"]]` standard deviations after transformation. The left panel shows SD ECDFs by group, and the right panel splits `r strings[["targets"]]` into below- and above-median abundance strata.

```{r sdecdf,fig.cap="(ref:sdecdf)", fig.height=6, fig.width=8}
p1 <- st$density(ggstat = "ecdf") +
  ggplot2::labs(title = "All quantified targets", tag = "A", x = "SD", y = "ECDF")
p2 <- st$density_median(ggstat = "ecdf") +
  ggplot2::labs(title = "Below and above median abundance", tag = "B", x = "SD", y = "ECDF")
prolfquapp::plotly_ggplot_subplot(p1, p2)
```

```{r sdVsMean, fig.cap="Standard deviation as a function of mean transformed intensity. Each point is one quantified target, the x-axis shows mean transformed intensity, and the y-axis shows SD.", eval=params$plot_sd_vs_mean}
st$stdv_vs_mean()
```


```{r computeSDQuantiles, include=FALSE}
sd_quantile_res2 <- st$stats_quantiles(probs = c(0.5, 0.6, 0.7, 0.8, 0.9))$wide
```

```{r printSDTable}
#knitr::kable(sd_quantile_res2,  digits = 3,
#             caption = "Summary of the distribution of standard deviations at the 50th, 60th, 70th, 80th and 90th percentile.", format="html") 
prolfqua::table_facade(sd_quantile_res2,
  caption = "Summary of the distribution of standard deviations at the 50th, 60th, 70th, 80th and 90th percentile.")

```

(ref:overviewHeat) Heatmap of transformed `r strings[["target"]]` intensities across samples. Rows are `r strings[["targets"]]`, columns are samples, and colours encode transformed intensity.

```{r prepareoverviewHeat, include = FALSE}

hm <- plotter_transformed$heatmap()
width = max(10, nrSamples * 0.15)
```

```{r overviewHeat, fig.cap="(ref:overviewHeat)", fig.height=10, fig.width=width}
print(hm)
```

# Sample Size Calculation


In the previous section, we estimated `r strings[["target"]]` variance from the QC samples. Figure \@ref(fig:sdviolinplots) shows the distribution of standard deviations after transformation. We use these SD estimates, together with typical significance and power settings, to estimate the sample sizes needed for the main experiment.

An important factor in estimating sample size is the smallest effect size you want to detect between two conditions, such as a reference and a treatment. Smaller biologically relevant effects require more samples. Typical $log_2$ fold-change thresholds are $0.59, 1, 2$, corresponding to fold changes of $1.5, 2, 4$.


Table \@ref(tab:sampleSize) and Figure \@ref(fig:figSampleSize) summarize how many samples are needed to detect $log_2$ fold-change differences of $0.59, 1, 2$ at a significance level of $5\%$ and power of $80\%$, using the SD quantiles for $50\%$ and $75\%$ of the measured `r strings[["targets"]]`.

(ref:figSampleSize) Estimated sample size for detecting $log_2$ fold-change differences by t-test. Bars show the number of samples required at significance level $0.05$ and power $0.8$ for the 50% and 75% SD quantiles of the measured `r strings[["targets"]]`; facets show the tested effect sizes.

```{r figSampleSize, fig.cap="(ref:figSampleSize)", fig.width=8, fig.alt=""}

sampleSize2 <- st$power_t_test_quantiles(probs = c(0.5,0.75), delta = c(0.59,1,2))
if (is.null(sampleSize2)) {
  message("Sample size estimation requires transformed data. Skipping.")
} else {
nudgeval <- max(sampleSize2$N) * 0.05

sampleSize2 <- sampleSize2 |> dplyr::mutate(delta = paste0("delta = ", delta))

#stats_res <- summarize_stats(dataTransformed$data_long(), dataTransformed$get_config())
#xx <- summarize_stats_quantiles(stats_res, local_config,probs = c(0.5,0.75))
#lfq_power_t_test_quantiles_V2(xx$long,delta = c(0.59,1,2))

sample_size_plot <- ggplot2::ggplot(sampleSize2, ggplot2::aes(x = probs, y = N)) +
  ggplot2::geom_bar(stat = "identity", color = "black", fill = "white") +
  ggplot2::geom_text(ggplot2::aes(label = N), nudge_y = nudgeval) +
  ggplot2::facet_wrap(as.formula(paste0("~ ", "delta + " , paste0(lfqdata$relevant_factor_keys(), collapse = " + "))))

tmp <- sampleSize2 |> tidyr::pivot_wider(id_cols = c("probs", "sdtrimmed", lfqdata$relevant_factor_keys()), names_from = delta, values_from = N)
print(sample_size_plot)
}

```


```{r sampleSize}
if (exists("tmp") && !is.null(tmp)) {
prolfqua::table_facade(tmp,
  caption = "Estimated sample size required to detect each tested log2 fold-change difference at significance level 0.05 and power 0.8 with a t-test.")
}

```


The _power_ of a test is $1-\beta$, where $\beta$ is the probability of a Type 2 error (failing to reject the null hypothesis when the alternative hypothesis is true).
In other words, if you have a $20\%$ chance of failing to detect a real difference, then the power of your test is $80\%$.

The _confidence level_ is equal to $1 - \alpha$, where $\alpha$ is the probability of making a Type 1 Error. That is, alpha represents the chance of a falsely rejecting $H_0$ and picking up a false-positive effect. Alpha is usually set at $5\%$ significance level, for a $95\%$ confidence level.

Fold change: Suppose you are comparing a treatment group to a placebo group, and you will be measuring some continuous response variable which, you hypothesize, will be affected by the treatment. We can consider the mean response in the treatment group, $\mu_1$, and the mean response in the placebo group, $\mu_2$. We can then define $\Delta = \mu_1 - \mu_2$ as the mean difference. The smaller the difference you want to detect, the larger the required sample size.


# Appendix

```{r sampleNameRawFileMapping}
prolfqua::table_facade(
  dataTransformed$factors(),
  caption = "Mapping of raw file names to sample names used throughout this report.")

```

```{r hierarchyCountsSample}
caption <- paste("Number of quantified ", strings[[".analyzed_target_types"]], " per sample.")
st <- dataTransformed$get_Summariser()
prolfqua::table_facade(
  st$hierarchy_counts_sample()
  , caption = caption)

```


```{r resetTheme, include=FALSE}
ggplot2::theme_set(old.theme)
```
