---
title: "Interaction Proteomics Report"
author: "Functional Genomics Center Zurich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
  bookdown::html_document2:
    toc: true
  pdf_document: 
    toc: true
params:
  sep: NULL
  FCt: 4
  SCt: 0.95 
---

```{r setup, include=FALSE}
library(DT)
library(ggraph)
library(igraph)
library(prolfqua)
library(tidyverse)
SEP <- params$sep
sp_string <- if (SEP$spc) {"spectral counts"} else {"intensities"}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

BFABRIC <- SEP$BFABRIC
```


# B-fabric related information

- Order ID: `r BFABRIC$orderID`

The input data, used to compile this report, is available through the b-fabric resource: [`r BFABRIC$inputID`](`r BFABRIC$inputURL`). The b-fabric dataset `r BFABRIC$datasetID` was used to annotate the LC-MS files. The complete output, including this report, R code used for analysis and intermediate results, is available through the b-fabric workunit: [`r BFABRIC$workunitID`](`r BFABRIC$workunitURL`).

# Introduction

- General introduction into interaction proteomics analysis
- Terminology: bait, prey
- AP-MS, Proximity labeling

# Results

- Overview of results
- Aim
- How to use the results

The LC-MS data was processed using the [FragPipe proteomics pipeline](https://fragpipe.nesvilab.org/). The protein quantification results were extracted from the _combined_protein.tsv_ file.  We used the columns with the `r if(SEP$spc){'Spectral Count' }else{'Intensity'}` suffix, which stores the `r if(SEP$spc){'total number of PSMs in support of the protein identification'} else {'normalized protein intensity using the sequences after razor assignment'}`. For more information about the _combined_protein.tsv_ file see [FragPipe output tutorial]( https://fragpipe.nesvilab.org/docs/tutorial_fragpipe_outputs.html#combined_proteintsv).

The table below lists the LC-MS files and their assigned annotations.

```{r sampleAnnotation}
DT::datatable(SEP$lfqdata_transformed$factors(),
              caption = "LC-MS raw file annotation table. The content of the sampleName column is used as a short form plot label and constructed from the CorT and bait attributes. CorT indicates if a raw file was assigned to the control group (C) or to the treatment group (T). The bait protein name is shown in the bait column. It should follow the <name>_<condition> pattern.")
```


## Peptide and Protein identification


In total `r SEP$lfqdata_transformed$hierarchy_counts()[2]` proteins (min. 2 peptides per protein) were identified and quantified using `r sp_string` across all samples (we will denote this as the protein abundance estimate from here on). Bar plot below shows the number of identified proteins per raw file.

```{r nrPerSample, fig.cap="Number of identified proteins across raw files.", fig.with=10, fig.height=7}
sum <- SEP$lfqdata_transformed$get_Summariser()

sum$plot_hierarchy_counts_sample()

```

### Protein abundance analysis

The density plot (Figure \@ref(fig:violinPlot)) displays the protein abundance distribution for all raw files of the data set. Mayor differences between raw files could be a hint that the individual protein abundance values are affected by technical biases. These biases might need to be corrected, in order to separate them from biological effects.

(ref:violinPlot) Protein abundance distribution. The dots represent the median log2-transformed abundance value per raw file. The overall density distribution is shown as a rotated kernel density plot.

```{r violinPlot, fig.cap="(ref:violinPlot)", fig.with=10, fig.height=7}
pl <- SEP$lfqdata_transformed$get_Plotter()
pl$intensity_distribution_density()
hm <- SEP$lfqdata_transformed$get_Plotter()$heatmap()

```


The protein abundance heatmap (Figure \@ref(fig:heatmap)) groups the protein and samples using unsupervised hierarchical clustering. Proteins and samples showing similar $\log_2$ abundances are grouped and shown in adjacent rows and columns respectively.

(ref:heatmap) Protein abundance heatmap (rows indicate proteins, columns indicate raw files) showing the row scaled $\log_2$ transformed protein abundance value. Co-clustering (hierarchical complete linkage, euclidean distance) of samples and proteins was used.

```{r heatmap, echo=FALSE, include=TRUE, fig.cap="(ref:heatmap)", fig.width=7, fig.height=7}
hm
```

We use principal component analysis (PCA) to transform the high dimensional space defined by all proteins into a two-dimensional one containing most of the information. Plot \@ref(fig:pca) shows the location of the samples according to the first and second principal component, which explain most of the variance in the data. Samples close in the PCA plot are more similar than those farther apart.

(ref:pca)  Plot of first and second principal component (PC1 and PC2) of principal component analysis (PCA). $\log_2$ transformed `r sp_string` are used as input.

```{r pca, fig.cap = "(ref:pca)" , fig.width=7, fig.height=7 }
SEP$lfqdata_transformed$get_Plotter()$pca_plotly()
```


### Missing value analysis

The analysis of missing values can be an important indicator for potential problems and biases in the data. We therefore visualize the structure of missing values (missing protein abundance estimate per protein) using different plots. Figure \@ref(fig:vennProteins) displays how many proteins are shared among the conditions or all possible intersection of conditions. If a protein was observed in one or zero samples in a group we say it is absent and present otherwise. We did choose zero ore one as absent, since one observation might be an artefact of a match between runs.


(ref:vennProteins) Venn diagram showing the number of proteins present in each group and in all possible intersections among groups. 

```{r vennProteins, fig.cap="(ref:vennProteins)", fig.with=10, fig.height=7, include = TRUE}
UpSetR::upset(SEP$pups, order.by = "freq", nsets = 1000)

```


```{r raster, echo=FALSE}
hm <- pl$NA_heatmap()
```

By transforming protein abundance estimates into present/absent calls a dichotomous view on the data (Figure \@ref(fig:naHeat) ) can be constructed.

(ref:naHeat) Protein abundance heatmap (rows indicate proteins, columns indicate raw files) showing missing protein abundance estimates across data set. Rows and columns are grouped based on the Minkowski distance using hierarchical clustering. White: Protein is observed, black: Protein is not observed.

```{r naHeat, echo=FALSE, fig.cap="(ref:naHeat)", fig.width=7, fig.height=7}
hm
```


## Interaction scoring using SAINT

In order to score potential interactions between observed proteins (potential prays) and the bait protein we used the [SaintExpress software](http://saint-apms.sourceforge.net/Main.html). The plots below show one subpanel for each bait present in the data set.

Please note, an empirical fold-change (EFC) score of e.g. 2 reported by SAINTexpress can not be interpreted as a two times higher protein abundance, because SAINTexpress internally normalizes and transforms the data.




The volcano plot \@ref(fig:volcanoPlotBfdr) helps to identify prey proteins with large EFC score and a low BFDR. The significance dimension is a $-\log_{10}$ transformed BFDR, i.e., small values of BFDR become large after transformation. Promising candidate pray proteins are found in the upper right sector of the plot.

(ref:volcanoPlotBfdr) Volcano plot showing $-\log_{10}$ transformed BFDR as function of $\log_2$ transformed effect size. The red line shows the $-log_{10}$ transformed BFDR threshold of `r SEP$FDRthreshold`, while the green line the $\log_2$ transformed FC of `r SEP$FCthreshold`. 




```{r setupVolcano, echo = FALSE}
datax <- SEP$resContrasts
prot_annot <- dplyr::rename(SEP$prot_annot, Prey=protein)
prot_annot <- dplyr::inner_join(prot_annot, datax)
bb <- prot_annot %>% dplyr::select(Prey, description, Bait, BFDR )
bb <- bb |> tidyr::pivot_wider(names_from = "Bait", names_prefix = "BFDR.", values_from = BFDR)
bb <- crosstalk::SharedData$new(bb, ~Prey, group = "Choose protein")
DT::datatable(bb, filter = "bottom", options = list(pageLength = 5))

```


```{r volcanoPlotBfdr, fig.cap = "(ref:volcanoPlotBfdr)", fig.width=9, fig.height=7, include = TRUE, eval = TRUE}
xd <- prolfqua::volcano_plotly( 
  datax ,
  effect = "log2_EFCs",
  significance = "BFDR",
  contrast = "Bait",
  xintercept =  log2(SEP$FCthreshold),
  yintercept = SEP$FDRthreshold,
  title_size = 10)

xd <- lapply(xd, plotly::highlight , off = "plotly_doubleclick")
nrow <- ceiling(length(xd) / 4)

plotly::subplot(xd, shareX = TRUE, shareY = TRUE, nrows = nrow)

```







## Interaction network reconstruction

The Bait Prey interaction candidate lists are filtered using an Bayesian false discovery rate (BFDR) threshold of __`r SEP$FDRthreshold`__ and an  empirical fold-change score (EFC) threshold of __`r SEP$FCthreshold`__ which corresponds to a $\log_2(EFC)$ of `r log2(SEP$FCthreshold)`.
Table \@ref(tab:nrSignificant) shows the number of bait prey pairs which pass the filter criteria, while \@ref(fig:SigPrey) lists the filtered bait prey pairs and shows their `BFDR` and `log2_EFCs` scores.


```{r nrSignificant}

acc <- SEP$sig |> dplyr::group_by(Bait) |> dplyr::summarize(accepted = dplyr::n())
all <- SEP$resContrasts |> dplyr::group_by(Bait) |> dplyr::summarise(all = dplyr::n())

accall <- dplyr::left_join(all, acc)
accall[is.na(accall)] <- 0
accall <- accall |> dplyr::mutate(`% accpeted ` = round(accepted / all * 100, digits = 1))

accall |>  knitr::kable( caption =
                 paste0("# of interactions accepted with BFDR < ", SEP$FDRthreshold," and EFC score threshold > ", SEP$FCthreshold,".")
                 )
isSig <- nrow(SEP$sig) > 0


```



```{r SigPrey, eval=isSig, fig.cap= "Candidate list."}
sig <- crosstalk::SharedData$new(SEP$sig, ~Prey, group = "Choose protein")
DT::datatable(sig)
```




```{r results='asis'}
if (isSig) {cat("The protein abundance heatmap (Figure \\@ref(fig:heatsig)) shows the proteins passing the BFDR and EFCs thresholds.")
} else {
  cat("There are not significant results given the _BFDR_ and EFC score thresholds.")
}

```

```{r aChunkNamePLEASE}
sigsub <- SEP$lfqdata_transformed$get_subset(data.frame(protein_Id = SEP$sig$Prey))
rs <- sigsub$get_Plotter()$raster(rownames = TRUE)

```


(ref:heatsig) Heatmap showing the $\log_2$ transformed protein abundance estimate for prays which pass the BFDR and EFCs thresholds.

```{r heatsig, fig.cap="(ref:heatsig)", fig.width=8, fig.height=10, eval = isSig}
rs
```



```{r intext , results='asis', eval=isSig}
cat("Protein interaction network reconstruction using graph. Nodes represent bait proteins are shown using the protein ID and bait proteins are shown using the name of the bait. Edges indicate bait-pray interactions scored above the chosen thresholds. Line color depends on the bait proteins.")
```


(ref:network) Protein interaction network for _BFDR_ < `r SEP$SSthreshold` and EFC score > `r SEP$FCthreshold`. Lines - protein interactions. Line color depends on the bait proteins.

```{r network, fig.cap="(ref:network)", eval=isSig}

bbSE <- dplyr::select(SEP$sig,
                      from = Bait,
                      to = Prey, 
                      Bait = Bait,
                      SS = SaintScore)

inn <- igraph::graph_from_data_frame( bbSE, directed = FALSE )

bb <- ggraph::ggraph(inn,layout = 'kk') +
  ggraph::geom_node_text(ggplot2::aes(label = name), size = 3.5) +
  ggraph::geom_edge_link(ggplot2::aes(color = Bait), alpha = 0.6) + 
  ggplot2::theme(legend.position = "none") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
bb

```


(ref:vennDiagram) Venn diagram showing the number of prays common and specific to the bait proteins in the data set.

```{r vennDiagram, fig.cap="(ref:vennDiagram)", eval = isSig}
b1 <- bbSE |> dplyr::select(from, to)

xx <-  split(b1$to, b1$from)

if (length(xx) > 1) {
  UpSetR::upset(UpSetR::fromList(xx), order.by = "freq", nsets = 1000)
}

```


# Additional Analysis


In the zip file you will find an excel file. All the figures in this document can be recreated using the data in this excel file.

The file contains several spreadsheets:
- _annotation_ - the annotation of the samples in the experiment
- _inter_, _prey_, and _bait_ spreadsheets which are the inputs to the SAINTexpress executable.
- _list_ spreadsheet which is the ouptut of SAINTExpress. 
- _MissingInformtion_ - spreadsheet containing information in how many samples of condition a protein is observed. This information is used to create the VennDiagram figure \@ref(fig:vennProteins).
- _inputData_ A table where each column represents a sample and each row represents a protein and the cells store the protein abundances.

A detailed description of the SAINTexpress inputs and outputs can be found in the _SAINTexpress.doc_ included in the zip file.


To compare the obtained results with known protein interactions we recommend the [string-db.org](https://string-db.org/), which is curated database of protein-protein interaction networks for a large variaty of organisms. String-db can also be used to perform functional enrichment analysis. To simplify the data upload to string-db we include text files containing the uniprot ids:

- `ORA_background.txt` all prey proteins.
- `ORA_Bait_<baitname>.txt` prey proteins accepted with the BFDR and EFC specified.





# Supplement

```{r fig.cap="x axis - Saint Score, y axis = -log10(BFDR)", eval = isSig}

ggplot2::ggplot( SEP$resContrasts, ggplot2::aes(x = SaintScore, y = -log10(BFDR) ) ) +
  ggplot2::geom_point() +
  ggplot2::geom_vline(xintercept = SEP$SSthreshold, color = 'red') +
  ggplot2::facet_wrap( ~ Bait )

```

# Session Info

```{r sessionInfo}
pander::pander(sessionInfo())
```

# References
