---
metadata-files:
  - _fgcz-report.yml
params:
  se_file: "../../extdata/3106962.rds"
  fdr_threshold: 0.05
  diff_threshold: 1
---

```{r setup}
#| include: false

# ── Load SE and extract all data upfront ─────────────────────────────────────
se <- readRDS(params$se_file)

# Sample annotations
col_data <- as.data.frame(SummarizedExperiment::colData(se))
col_data$sampleName <- rownames(col_data)

# Detect group column
skip_cols <- c("sampleName", "Name", "raw.file", "CONTROL", "isotopeLabel")
group_col <- setdiff(colnames(col_data), skip_cols)[1]
groups <- sort(unique(col_data[[group_col]]))

# Assay matrices
mat_raw <- SummarizedExperiment::assay(se, "rawData")
mat_trans <- SummarizedExperiment::assay(se, "transformedData")

# Contrast results → combined long tibble
rd <- SummarizedExperiment::rowData(se)
contrast_names <- grep("^constrast_", names(rd), value = TRUE)
ctr_all <- dplyr::bind_rows(lapply(contrast_names, function(cn) {
  df <- as.data.frame(rd[[cn]])
  if (!"protein_Id" %in% colnames(df)) df$protein_Id <- rownames(df)
  if (!"contrast" %in% colnames(df)) df$contrast <- gsub("^constrast_", "", cn)
  df
}))

# Stats tables
stats_raw <- tibble::as_tibble(as.data.frame(rd[["stats_raw_wide"]]))
stats_norm <- tibble::as_tibble(as.data.frame(rd[["stats_normalized_wide"]]))

# Plots directory
dir.create("plots", showWarnings = FALSE)

# Consistent group colour palette
group_pal <- setNames(
  scales::hue_pal()(length(groups)),
  groups
)

# DT helper — scrollY auto-sizes for small tables
.dt <- function(df, ...) {
  sy <- min(300, max(80, nrow(df) * 35 + 40))
  DT::datatable(df, filter = "bottom", extensions = c("Scroller", "Buttons"),
    class = "compact", options = list(
      deferRender = TRUE, scrollY = sy, scrollX = TRUE, scroller = TRUE,
      dom = "Bfrtip", buttons = list("csv", "excel")
    ), ...)
}
```


::: {.panel-tabset}

## Missing Value Analysis

::: {.callout-note collapse="true"}
## About this plot
The left panels show the number of proteins with $0$–$N$ missing values per group.
The right panels show the distribution of mean protein intensity given the missingness level.
Proteins with zero missing values typically have higher average abundance.
:::

```{r}
#| label: fig-missingProtein
#| fig-cap: "Top: number of proteins with n missing values per group. Bottom: mean intensity distribution by missingness level."
#| fig-height: 8
#| fig-width: 9
#| lightbox: true

# Compute missingness per group
miss_df <- dplyr::bind_rows(lapply(groups, function(g) {
  samples <- col_data$sampleName[col_data[[group_col]] == g]
  sub <- mat_raw[, samples, drop = FALSE]
  data.frame(
    protein_Id = rownames(sub),
    nrNA = rowSums(is.na(sub)),
    group = g,
    meanAbundance = rowMeans(sub, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}))

p1 <- ggplot2::ggplot(
  miss_df |> dplyr::filter(.data$nrNA > 0),
  ggplot2::aes(x = factor(.data$nrNA))
) +
  ggplot2::geom_bar() +
  ggplot2::facet_wrap(~group) +
  ggplot2::labs(x = "Number of missing values (nrNA)", y = "Number of proteins") +
  ggplot2::theme_classic()

p2 <- ggplot2::ggplot(
  miss_df,
  ggplot2::aes(x = .data$meanAbundance, color = factor(.data$nrNA))
) +
  ggplot2::geom_density() +
  ggplot2::facet_wrap(~group) +
  ggplot2::labs(x = "Mean abundance", y = "Density", color = "nrNA") +
  ggplot2::theme_classic()

gridExtra::grid.arrange(p1, p2, nrow = 2)

ggplot2::ggsave("plots/missingness_analysis.pdf",
  gridExtra::arrangeGrob(p1, p2, nrow = 2), width = 9, height = 8)
```


## Variance Analysis

::: {.callout-note collapse="true"}
## About these plots
- **Panel A**: Coefficient of variation (CV) within each group and across all samples, computed on raw data. Ideally, within-group CV is smaller than overall CV.
- **Panel B**: Standard deviation (sd) on raw data.
- **Panel C**: Standard deviation (sd) on normalized data. Within-group variance should decrease after normalization.
:::

```{r}
#| label: fig-SDViolin
#| fig-cap: "A: CV of raw data. B: SD of raw data. C: SD of normalized data."
#| fig-height: 4
#| fig-width: 10
#| lightbox: true

# Panel A: CV from stats_raw_wide
cv_cols <- grep("^CV_", colnames(stats_raw), value = TRUE)
cv_long <- stats_raw |>
  dplyr::select("protein_Id", dplyr::all_of(cv_cols)) |>
  tidyr::pivot_longer(cols = -"protein_Id", names_to = "group", values_to = "CV") |>
  dplyr::mutate(group = gsub("^CV_", "", .data$group)) |>
  dplyr::filter(is.finite(.data$CV))

pA <- ggplot2::ggplot(cv_long, ggplot2::aes(x = .data$group, y = .data$CV)) +
  ggplot2::geom_violin(fill = "lightblue", alpha = 0.5) +
  ggplot2::stat_summary(fun = median, geom = "point", size = 2) +
  ggplot2::labs(x = "", y = "CV", tag = "A") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

# Panel B: SD from stats_raw_wide
sd_raw_cols <- grep("^sd_", colnames(stats_raw), value = TRUE)
sd_raw_long <- stats_raw |>
  dplyr::select("protein_Id", dplyr::all_of(sd_raw_cols)) |>
  tidyr::pivot_longer(cols = -"protein_Id", names_to = "group", values_to = "sd") |>
  dplyr::mutate(group = gsub("^sd_", "", .data$group)) |>
  dplyr::filter(is.finite(.data$sd))

pB <- ggplot2::ggplot(sd_raw_long, ggplot2::aes(x = .data$group, y = .data$sd)) +
  ggplot2::geom_violin(fill = "lightblue", alpha = 0.5) +
  ggplot2::stat_summary(fun = median, geom = "point", size = 2) +
  ggplot2::labs(x = "", y = "sd", tag = "B") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

# Panel C: SD from stats_normalized_wide
sd_norm_cols <- grep("^sd_", colnames(stats_norm), value = TRUE)
sd_norm_long <- stats_norm |>
  dplyr::select("protein_Id", dplyr::all_of(sd_norm_cols)) |>
  tidyr::pivot_longer(cols = -"protein_Id", names_to = "group", values_to = "sd") |>
  dplyr::mutate(group = gsub("^sd_", "", .data$group)) |>
  dplyr::filter(is.finite(.data$sd))

pC <- ggplot2::ggplot(sd_norm_long, ggplot2::aes(x = .data$group, y = .data$sd)) +
  ggplot2::geom_violin(fill = "lightblue", alpha = 0.5) +
  ggplot2::stat_summary(fun = median, geom = "point", size = 2) +
  ggplot2::labs(x = "", y = "sd", tag = "C") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

gridExtra::grid.arrange(pA, pB, pC, nrow = 1)

ggplot2::ggsave("plots/variance_analysis.pdf",
  gridExtra::arrangeGrob(pA, pB, pC, nrow = 1), width = 10, height = 4)
```

```{r}
#| label: tbl-CVtable
rows <- list()

if (length(cv_cols) > 0) {
  med_cv <- sapply(stats_raw[cv_cols], median, na.rm = TRUE)
  rows[["CV"]] <- data.frame(what = "CV", t(round(med_cv, 4)), check.names = FALSE)
  colnames(rows[["CV"]])[-1] <- gsub("^CV_", "", names(med_cv))
}
if (length(sd_raw_cols) > 0) {
  med_sd_raw <- sapply(stats_raw[sd_raw_cols], median, na.rm = TRUE)
  rows[["sd_raw"]] <- data.frame(what = "sd_raw", t(round(med_sd_raw, 4)), check.names = FALSE)
  colnames(rows[["sd_raw"]])[-1] <- gsub("^sd_", "", names(med_sd_raw))
}
if (length(sd_norm_cols) > 0) {
  med_sd_norm <- sapply(stats_norm[sd_norm_cols], median, na.rm = TRUE)
  rows[["sd_norm"]] <- data.frame(what = "sd_norm", t(round(med_sd_norm, 4)), check.names = FALSE)
  colnames(rows[["sd_norm"]])[-1] <- gsub("^sd_", "", names(med_sd_norm))
}

if (length(rows) > 0) {
  .dt(dplyr::bind_rows(rows))
}
```


## Fold-Change Distribution

::: {.callout-note collapse="true"}
## About this plot
Distribution of differences (log2 fold-change) between groups. The red dashed line marks the median, the green line the expected value (zero). If the median is non-zero, consider this when interpreting results.
:::

```{r}
#| label: fig-densityOFFoldChanges
#| fig-cap: "Fold-change distribution per contrast."
#| fig-width: 9
#| fig-height: 5
#| lightbox: true
ggplot2::ggplot(ctr_all, ggplot2::aes(x = .data$diff)) +
  ggplot2::geom_histogram(bins = 50, fill = "grey70", color = "black") +
  ggplot2::geom_vline(xintercept = 0, color = "green") +
  ggplot2::geom_vline(
    data = ctr_all |>
      dplyr::group_by(.data$contrast) |>
      dplyr::summarise(med = median(.data$diff, na.rm = TRUE), .groups = "drop"),
    ggplot2::aes(xintercept = .data$med),
    color = "red", linetype = "dashed"
  ) +
  ggplot2::facet_wrap(~contrast) +
  ggplot2::labs(x = "Difference (log2 FC)", y = "Count") +
  ggplot2::theme_classic()

ggplot2::ggsave("plots/fc_histogram.pdf", width = 9, height = 5)
```


## P-value Distribution

::: {.callout-note collapse="true"}
## About this plot
Under the null hypothesis the p-value distribution should be uniform. Enrichment of small p-values indicates true differential expression.
:::

```{r}
#| label: fig-pvalueHist
#| fig-cap: "P-value histogram per contrast."
#| fig-width: 9
#| fig-height: 5
#| lightbox: true
ggplot2::ggplot(ctr_all, ggplot2::aes(x = .data$p.value)) +
  ggplot2::geom_histogram(bins = 50, fill = "grey70", color = "black") +
  ggplot2::facet_wrap(~contrast) +
  ggplot2::labs(x = "p-value", y = "Count") +
  ggplot2::theme_classic()

ggplot2::ggsave("plots/pvalue_histogram.pdf", width = 9, height = 5)
```


## MA Plot

::: {.callout-note collapse="true"}
## About the MA plot
The MA plot shows differences between groups (y-axis) as a function of average protein abundance (x-axis).
The fold change should not depend on protein abundance.
Panel A shows abundance on the original scale; Panel B shows the rank of abundance.
:::

```{r}
#| label: fig-MAPlot
#| fig-cap: "MA plots. A: diff vs average abundance. B: diff vs rank of abundance."
#| fig-width: 9
#| fig-height: 8
#| lightbox: true
ppA <- ggplot2::ggplot(ctr_all, ggplot2::aes(x = .data$avgAbd, y = .data$diff)) +
  ggplot2::geom_point(alpha = 0.3, size = 0.5) +
  ggplot2::geom_hline(yintercept = c(-params$diff_threshold, params$diff_threshold),
                      linetype = "dashed", color = "green") +
  ggplot2::geom_hline(yintercept = 0, color = "grey50") +
  ggplot2::facet_wrap(~contrast) +
  ggplot2::labs(x = "Average abundance", y = "diff (log2 FC)", tag = "A") +
  ggplot2::theme_classic()

ctr_ranked <- ctr_all |>
  dplyr::group_by(.data$contrast) |>
  dplyr::mutate(rank_abd = rank(.data$avgAbd, na.last = "keep")) |>
  dplyr::ungroup()

ppB <- ggplot2::ggplot(ctr_ranked, ggplot2::aes(x = .data$rank_abd, y = .data$diff)) +
  ggplot2::geom_point(alpha = 0.3, size = 0.5) +
  ggplot2::geom_hline(yintercept = c(-params$diff_threshold, params$diff_threshold),
                      linetype = "dashed", color = "green") +
  ggplot2::geom_hline(yintercept = 0, color = "grey50") +
  ggplot2::facet_wrap(~contrast) +
  ggplot2::labs(x = "Rank of average abundance", y = "diff (log2 FC)", tag = "B") +
  ggplot2::theme_classic()

gridExtra::grid.arrange(ppA, ppB, ncol = 1)

ggplot2::ggsave("plots/ma_plots.pdf",
  gridExtra::arrangeGrob(ppA, ppB, ncol = 1), width = 9, height = 8)
```


## Session Info

```{r}
pander::pander(sessionInfo())
```

:::

---

*Generated from `DiffExpQC_SE.qmd` in the `prolfquapp` R package.*
