---
title: "Differential Expression Analysis"
metadata-files:
  - _fgcz-report.yml
params:
  se_file: "../../extdata/3106962.rds"
  fdr_threshold: 0.05
  diff_threshold: 1
---

```{r setup}
#| include: false
library(DT)
library(ggplot2)
library(gridExtra)
library(prolfqua)
library(prolfquapp)

report <- prolfquapp:::se_report_lfqdata(params$se_file)
se <- report$se
meta <- report$metadata
lfq_raw <- report$lfq_raw
lfq_transformed <- report$lfq_transformed
contrast_table <- report$contrast_table
contrast_object <- report$contrast_object

dir.create("plots", showWarnings = FALSE, recursive = TRUE)

empty_report_plot <- function(label) {
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = label) +
    ggplot2::theme_void()
}

safe_plot <- function(expr, label) {
  tryCatch(expr, error = function(e) {
    empty_report_plot(paste(label, conditionMessage(e), sep = "\n"))
  })
}

draw_or_empty <- function(expr, label) {
  tryCatch(
    expr,
    error = function(e) {
      show_plot(empty_report_plot(paste(label, conditionMessage(e), sep = "\n")))
    }
  )
}

show_plot <- function(plot) {
  if (is.null(plot)) {
    plot <- empty_report_plot("No plot is available for this data set.")
  }
  if (inherits(plot, "pheatmap")) {
    grid::grid.draw(plot$gtable)
  } else {
    print(plot)
  }
  invisible(plot)
}

save_if_ggplot <- function(plot, name, width = 8, height = 5) {
  if (inherits(plot, "ggplot")) {
    ggplot2::ggsave(file.path("plots", paste0(name, ".png")), plot, width = width, height = height, dpi = 300)
    ggplot2::ggsave(file.path("plots", paste0(name, ".pdf")), plot, width = width, height = height)
  }
  invisible(plot)
}

html_escape <- function(x) {
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x <- gsub("'", "&#39;", x, fixed = TRUE)
  x
}

html_link <- function(url) {
  sprintf(
    '<a href="%s" target="_blank" rel="noopener noreferrer">%s</a>',
    html_escape(url),
    html_escape(url)
  )
}

dt_table <- function(df, caption = NULL, page_length = 10, escape = TRUE) {
  if (is.null(df) || nrow(df) == 0) {
    df <- data.frame(message = "No rows available.")
  }
  DT::datatable(
    df,
    caption = caption,
    filter = "bottom",
    extensions = "Buttons",
    rownames = FALSE,
    class = "compact stripe",
    escape = escape,
    options = list(pageLength = page_length, scrollX = TRUE, dom = "Blfrtip", buttons = c("csv", "excel"))
  )
}

contrast_plotter <- NULL
if (!is.null(contrast_object)) {
  contrast_plotter <- contrast_object$get_Plotter(
    fc_threshold = params$diff_threshold,
    fdr_threshold = params$fdr_threshold
  )
}

significant_contrasts <- if (nrow(contrast_table) > 0 &&
    all(c("FDR", "diff") %in% colnames(contrast_table))) {
  contrast_table |>
    dplyr::filter(.data$FDR < params$fdr_threshold, abs(.data$diff) > params$diff_threshold)
} else {
  tibble::tibble()
}
```

::: {.panel-tabset}

# Settings

::: {.callout-note collapse="true"}
## Input

The report was generated from `r basename(params$se_file)`. It summarizes raw and transformed
protein abundance data together with differential expression results.
:::

```{r}
#| label: tbl-bfabric
urls <- meta$bfabric_urls
url_df <- data.frame(field = character(), url = character())
if (is.list(urls) && length(urls) > 0) {
  url_values <- vapply(urls, function(x) {
    if (is.null(x) || length(x) == 0 || all(is.na(x))) {
      return(NA_character_)
    }
    as.character(x[[1]])
  }, character(1))
  url_df <- data.frame(field = names(url_values), url = unname(url_values), stringsAsFactors = FALSE)
}
url_df <- url_df[!is.na(url_df$url) & nzchar(url_df$url), , drop = FALSE]
if (nrow(url_df) > 0) {
  url_df$field <- html_escape(url_df$field)
  url_df$url <- html_link(url_df$url)
}
dt_table(url_df, caption = "B-Fabric links", page_length = 5, escape = FALSE)
```

```{r}
#| label: tbl-analysis-settings
settings <- data.frame(
  parameter = c("FDR threshold", "Difference threshold", "Model formula"),
  value = c(
    params$fdr_threshold,
    params$diff_threshold,
    if (!is.null(meta$formula$formula)) meta$formula$formula[[1]] else NA
  )
)
dt_table(settings, caption = "Analysis settings", page_length = 5)
```

# Results

```{r}
#| label: tbl-sample-counts
sample_counts <- as.data.frame(table(lfq_raw$factors()[lfq_raw$relevant_factor_keys()]))
colnames(sample_counts)[ncol(sample_counts)] <- "samples"
dt_table(sample_counts, caption = "Number of samples per condition")
```

```{r}
#| label: tbl-significant-summary
if (nrow(significant_contrasts) > 0) {
  signif_summary <- significant_contrasts |>
    dplyr::mutate(direction = ifelse(.data$diff > 0, "up", "down")) |>
    dplyr::count(.data$contrast, .data$direction, name = "proteins")
} else {
  signif_summary <- tibble::tibble()
}
dt_table(signif_summary, caption = "Significant proteins by contrast and direction")
```

# Protein Identification

```{r}
#| label: fig-protein-counts
#| fig-cap: "Number of quantified proteins per sample."
#| fig-width: 10
#| fig-height: 6
protein_counts <- safe_plot(
  lfq_raw$get_Summariser()$plot_hierarchy_counts_sample(nr_children = 1),
  "Could not draw protein counts per sample."
)
save_if_ggplot(protein_counts, "protein_counts_per_sample", width = 10, height = 6)
show_plot(protein_counts)
```

```{r}
#| label: fig-protein-counts-two-peptides
#| fig-cap: "Number of quantified proteins with at least two peptides per sample."
#| fig-width: 10
#| fig-height: 6
protein_counts_two_peptides <- safe_plot(
  lfq_raw$get_Summariser()$plot_hierarchy_counts_sample(nr_children = 2),
  "Could not draw protein counts with at least two peptides per sample."
)
if (is.null(protein_counts_two_peptides)) {
  protein_counts_two_peptides <- empty_report_plot(
    "No proteins with at least two peptides are available for this data set."
  )
}
save_if_ggplot(protein_counts_two_peptides, "protein_counts_two_peptides_per_sample", width = 10, height = 6)
show_plot(protein_counts_two_peptides)
```

```{r}
#| label: tbl-sample-annotation
sample_annotation <- lfq_raw$factors()
dt_table(sample_annotation, caption = "Sample annotation and experimental factors")
```

# Quality Control

::: {.panel-tabset}

## Missing Values

```{r}
#| label: fig-missing-heatmap
#| fig-cap: "Missing values in raw protein abundances."
#| fig-width: 7
#| fig-height: 6
show_plot(safe_plot(
  lfq_raw$get_Plotter()$na_heatmap(),
  "Could not draw missing value heatmap."
))
```

```{r}
#| label: fig-missingness-per-group
#| fig-cap: "Missingness summaries for raw protein abundances."
#| fig-width: 15
#| fig-height: 4.5
missing_per_group <- safe_plot(
  lfq_raw$get_Summariser()$plot_missingness_per_group(),
  "Could not draw missingness per group."
)
missing_cumsum <- safe_plot(
  lfq_raw$get_Summariser()$plot_missingness_per_group_cumsum(),
  "Could not draw cumulative missingness per group."
)
missing_hist <- safe_plot(
  lfq_raw$get_Plotter()$missigness_histogram(),
  "Could not draw missingness intensity histogram."
)
combined_missing <- gridExtra::arrangeGrob(missing_per_group, missing_cumsum, missing_hist, ncol = 3)
save_if_ggplot(missing_per_group, "missingness_per_group", width = 9, height = 4)
save_if_ggplot(missing_cumsum, "missingness_per_group_cumsum", width = 9, height = 4)
save_if_ggplot(missing_hist, "missingness_histogram", width = 9, height = 4)
grid::grid.draw(combined_missing)
```

```{r}
#| label: fig-protein-detection-overlap
#| fig-cap: "Protein detections shared across experimental groups."
#| fig-width: 9
#| fig-height: 6
draw_or_empty(
  lfq_raw$get_Summariser()$upset_interaction_missing_stats(tr = 1),
  "Could not draw protein detection overlap plot."
)
```

## Abundance Distributions

```{r}
#| label: fig-abundance-density
#| fig-cap: "Raw and transformed abundance distributions."
#| fig-width: 10
#| fig-height: 5
raw_density <- safe_plot(
  lfq_raw$get_Plotter()$intensity_distribution_density(),
  "Could not draw raw abundance density."
)
trans_density <- safe_plot(
  lfq_transformed$get_Plotter()$intensity_distribution_density(),
  "Could not draw transformed abundance density."
)
save_if_ggplot(raw_density, "raw_abundance_density")
save_if_ggplot(trans_density, "transformed_abundance_density")
gridExtra::grid.arrange(raw_density, trans_density, nrow = 1)
```

## Variance

```{r}
#| label: fig-variance-violin
#| fig-cap: "Raw protein abundance CV, log2 raw protein abundance SD, and transformed protein abundance SD distributions."
#| fig-width: 15
#| fig-height: 5
raw_violin <- safe_plot(
  lfq_raw$get_Stats()$violin(),
  "Could not draw raw CV violin plot."
)
log2_raw <- lfq_raw$get_Transformer()$log2()$lfq
log2_raw_violin <- safe_plot(
  log2_raw$get_Stats()$violin(),
  "Could not draw log2 raw SD violin plot."
)
trans_violin <- safe_plot(
  lfq_transformed$get_Stats()$violin(),
  "Could not draw transformed SD violin plot."
)
save_if_ggplot(raw_violin, "raw_cv_violin")
save_if_ggplot(log2_raw_violin, "log2_raw_sd_violin")
save_if_ggplot(trans_violin, "transformed_sd_violin")
gridExtra::grid.arrange(raw_violin, log2_raw_violin, trans_violin, nrow = 1)
```

```{r}
#| label: tbl-variance-summary
variance_summary <- dplyr::bind_rows(
  raw_cv = lfq_raw$get_Stats()$stats_quantiles()$wide,
  transformed_sd = lfq_transformed$get_Stats()$stats_quantiles()$wide,
  .id = "source"
)
dt_table(variance_summary, caption = "Variance quantile summary")
```

## Sample Structure

::: {.panel-tabset}

### PCA

```{r}
#| label: fig-pca
#| fig-cap: "PCA of transformed protein abundances."
#| fig-width: 7
#| fig-height: 6
pca_plot <- safe_plot(
  lfq_transformed$get_Plotter()$pca(),
  "Could not draw PCA."
)
save_if_ggplot(pca_plot, "pca", width = 7, height = 6)
show_plot(pca_plot)
```

### Correlation

```{r}
#| label: fig-correlation
#| fig-cap: "Sample correlation heatmap for transformed protein abundances."
#| fig-width: 7
#| fig-height: 6
show_plot(safe_plot(
  lfq_transformed$get_Plotter()$heatmap_cor(),
  "Could not draw sample correlation heatmap."
))
```

### Abundance Heatmap

```{r}
#| label: fig-abundance-heatmap
#| fig-cap: "Heatmap of transformed protein abundances."
#| fig-width: 7
#| fig-height: 6
show_plot(safe_plot(
  lfq_transformed$get_Plotter()$heatmap(),
  "Could not draw abundance heatmap."
))
```

:::

:::

# Differential Expression

::: {.panel-tabset}

## Overview

```{r}
#| label: fig-fc-pvalue
#| fig-cap: "Fold-change and p-value summaries."
#| fig-width: 10
#| fig-height: 5
if (!is.null(contrast_plotter)) {
  fc_hist <- safe_plot(contrast_plotter$histogram_diff(), "Could not draw fold-change histogram.")
  p_hist <- safe_plot(contrast_plotter$histogram()$p.value, "Could not draw p-value histogram.")
  save_if_ggplot(fc_hist, "fold_change_histogram")
  save_if_ggplot(p_hist, "pvalue_histogram")
  gridExtra::grid.arrange(fc_hist, p_hist, nrow = 1)
} else {
  show_plot(empty_report_plot("No contrast results are available."))
}
```

## Volcano

```{r}
#| label: fig-volcano
#| fig-cap: "Differential expression volcano plot."
#| fig-width: 9
#| fig-height: 7
if (!is.null(contrast_plotter)) {
  volcano_plot <- safe_plot(contrast_plotter$volcano()$FDR, "Could not draw volcano plot.")
  save_if_ggplot(volcano_plot, "volcano", width = 9, height = 7)
  show_plot(volcano_plot)
} else {
  show_plot(empty_report_plot("No contrast results are available."))
}
```

## MA Plot

```{r}
#| label: fig-ma
#| fig-cap: "Differential expression MA plots by average abundance and abundance rank."
#| fig-width: 8
#| fig-height: 9
if (!is.null(contrast_plotter)) {
  ma_plot <- safe_plot(contrast_plotter$ma_plot(rank = FALSE), "Could not draw MA plot.")
  ma_rank_plot <- safe_plot(
    contrast_plotter$ma_plot(fc = params$diff_threshold, rank = TRUE),
    "Could not draw ranked MA plot."
  )
  save_if_ggplot(ma_plot, "ma_plot", width = 8, height = 5)
  save_if_ggplot(ma_rank_plot, "ma_rank_plot", width = 8, height = 5)
  gridExtra::grid.arrange(ma_plot, ma_rank_plot, ncol = 1)
} else {
  show_plot(empty_report_plot("No contrast results are available."))
}
```

## Significant Protein Heatmap

```{r}
#| label: fig-significant-protein-heatmap
#| fig-cap: "Transformed protein abundances for significant differential-expression results."
#| fig-width: 9
#| fig-height: 8
if (nrow(significant_contrasts) > 0 && "protein_Id" %in% colnames(significant_contrasts)) {
  significant_lfq <- lfq_transformed$get_copy()
  significant_lfq <- significant_lfq$get_subset(significant_contrasts)
  row_names <- tryCatch(significant_lfq$hierarchy_counts()[[2]] <= 30, error = function(e) FALSE)
  significant_heatmap <- safe_plot(
    significant_lfq$get_Plotter()$raster(rownames = row_names),
    "Could not draw significant protein heatmap."
  )
  show_plot(significant_heatmap)
} else {
  show_plot(empty_report_plot("No significant differential-expression results are available."))
}
```

## Significant Protein Overlap

```{r}
#| label: fig-significant-protein-overlap
#| fig-cap: "Significant proteins shared across differential-expression contrasts."
#| fig-width: 9
#| fig-height: 6
if (nrow(significant_contrasts) > 0 &&
    all(c("protein_Id", "contrast") %in% colnames(significant_contrasts))) {
  significant_sets <- split(significant_contrasts[["protein_Id"]], significant_contrasts[["contrast"]])
  significant_sets <- significant_sets[lengths(significant_sets) > 0]
  if (length(significant_sets) > 1) {
    draw_or_empty(
      UpSetR::upset(UpSetR::fromList(significant_sets), order.by = "freq"),
      "Could not draw significant protein overlap plot."
    )
  } else {
    show_plot(empty_report_plot("Significant proteins are available for fewer than two contrasts."))
  }
} else {
  show_plot(empty_report_plot("No significant differential-expression results are available."))
}
```

:::

# Tables

```{r}
#| label: tbl-contrast-results
dt_table(contrast_table, caption = "Differential expression results", page_length = 25)
```

# Session Info

```{r}
sessionInfo()
```

:::
