---
metadata-files:
  - _fgcz-report.yml
params:
  se_file: "../../extdata/3106962.rds"
  fdr_threshold: 0.05
  diff_threshold: 1
---

```{r setup}
#| include: false

# ── Load SE and extract all data upfront ─────────────────────────────────────
se <- readRDS(params$se_file)

# Sample annotations
col_data <- as.data.frame(SummarizedExperiment::colData(se))
col_data$sampleName <- rownames(col_data)

# Detect group column (first that isn't sampleName/Name/raw.file/CONTROL/isotopeLabel)
skip_cols <- c("sampleName", "Name", "raw.file", "CONTROL", "isotopeLabel")
group_col <- setdiff(colnames(col_data), skip_cols)[1]

# Assay matrices
mat_raw <- SummarizedExperiment::assay(se, "rawData")
mat_trans <- SummarizedExperiment::assay(se, "transformedData")

# Assays → long format
.assay_to_long <- function(mat, col_data, group_col) {
  df <- as.data.frame(mat, check.names = FALSE)
  df$protein_Id <- rownames(mat)
  long <- tidyr::pivot_longer(df, cols = -"protein_Id",
                               names_to = "sampleName", values_to = "abundance")
  dplyr::left_join(long, col_data, by = "sampleName")
}
long_raw <- .assay_to_long(mat_raw, col_data, group_col)
long_trans <- .assay_to_long(mat_trans, col_data, group_col)

# Contrast results → combined long tibble
rd <- SummarizedExperiment::rowData(se)
contrast_names <- grep("^constrast_", names(rd), value = TRUE)
ctr_all <- dplyr::bind_rows(lapply(contrast_names, function(cn) {
  df <- as.data.frame(rd[[cn]])
  if (!"protein_Id" %in% colnames(df)) df$protein_Id <- rownames(df)
  if (!"contrast" %in% colnames(df)) df$contrast <- gsub("^constrast_", "", cn)
  df
}))

# Significant
ctr_signif <- ctr_all |>
  dplyr::filter(.data$FDR < params$fdr_threshold,
                abs(.data$diff) > params$diff_threshold)

# Stats tables
stats_raw <- tibble::as_tibble(as.data.frame(rd[["stats_raw_wide"]]))
stats_norm <- tibble::as_tibble(as.data.frame(rd[["stats_normalized_wide"]]))

# Metadata
meta <- S4Vectors::metadata(se)

# Summary counts
ids <- rownames(se)
n_total <- length(ids)
n_con <- sum(grepl("^zz|^CON", ids))
n_rev <- sum(grepl("^REV|^rev", ids))
pct_con <- round(n_con / n_total * 100, 2)
pct_rev <- round(n_rev / n_total * 100, 2)

# Protein counts per sample
n_proteins_per_sample <- colSums(!is.na(mat_raw))

# Plots directory
dir.create("plots", showWarnings = FALSE)

# Consistent group colour palette used across all figures
groups <- sort(unique(col_data[[group_col]]))
group_pal <- setNames(
  scales::hue_pal()(length(groups)),
  groups
)

# DT helper for consistent table styling
# scrollY auto-sizes: small tables get compact height, large tables scroll
.dt <- function(df, ...) {
  sy <- min(300, max(80, nrow(df) * 35 + 40))
  DT::datatable(df, filter = "bottom", extensions = c("Scroller", "Buttons"),
    class = "compact", options = list(
      deferRender = TRUE, scrollY = sy, scrollX = TRUE, scroller = TRUE,
      dom = "Bfrtip", buttons = list("csv", "excel")
    ), ...)
}
```


::: {.panel-tabset}

## B-fabric Info

::: {.callout-note}
## Project Information

```{r}
#| results: asis
urls <- meta$bfabric_urls
if (!is.null(urls$projectURL))
  cat(sprintf("- **Project**: [%s](%s)\n", basename(urls$projectURL), urls$projectURL))
if (!is.null(urls$orderURL))
  cat(sprintf("- **Order**: [%s](%s)\n", basename(urls$orderURL), urls$orderURL))
if (!is.null(urls$workunitURL))
  cat(sprintf("- **Workunit**: [%s](%s)\n", basename(urls$workunitURL), urls$workunitURL))
cat(sprintf("\n**Model formula**: `%s`\n", meta$formula$formula[1]))
```

:::


## Results

```{r}
#| label: tbl-samples
tab <- as.data.frame(table(col_data[[group_col]]))
colnames(tab) <- c(group_col, "# samples")
.dt(tab)
```

```{r}
#| label: tbl-annotation
ann <- col_data
ann$n_proteins <- n_proteins_per_sample[ann$sampleName]
.dt(ann)
```


## Protein Identification

::: {.callout-note}
## Summary

- Total proteins (excl. decoys): **`r n_total - n_rev`**
- Contaminant proteins: **`r pct_con`%**
- False positives (decoys): **`r pct_rev`%**
:::

```{r}
#| label: fig-nrPerSample
#| fig-cap: "Number of quantified proteins per sample."
#| fig-width: 10
#| fig-height: 6
#| lightbox: true
pc_df <- data.frame(
  sampleName = names(n_proteins_per_sample),
  n_proteins = n_proteins_per_sample
)
pc_df <- dplyr::left_join(pc_df, col_data, by = "sampleName")

ggplot2::ggplot(pc_df, ggplot2::aes(
  x = .data$sampleName, y = .data$n_proteins, fill = .data[[group_col]]
)) +
  ggplot2::geom_col() +
  ggplot2::scale_fill_manual(values = group_pal) +
  ggplot2::labs(x = "", y = "Number of proteins") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

ggplot2::ggsave("plots/protein_counts_per_sample.pdf", width = 10, height = 6)
```


## Missing Value Analysis

```{r}
#| include: false
na_mat <- ifelse(is.na(mat_raw), 1, 0)
na_mat_subset <- na_mat[rowSums(na_mat) > 0, , drop = FALSE]
n_with_na <- nrow(na_mat_subset)
pct_with_na <- round(n_with_na / nrow(mat_raw) * 100, 1)
```

There are **`r n_with_na`** proteins with at least one missing value (**`r pct_with_na`%**).

::: {.panel-tabset}

### Heatmap

::: {.callout-note collapse="true"}
## About this figure
Protein abundance heatmap showing missing values. White = observed, Black = missing.
Rows and columns clustered using hierarchical clustering.
:::

```{r}
#| label: fig-naHeat
#| fig-cap: "Missing value heatmap (black = missing, white = present)."
#| fig-width: 7
#| fig-height: 6
#| lightbox: true
if (nrow(na_mat_subset) > 0) {
  col_ann <- ComplexHeatmap::HeatmapAnnotation(
    group = col_data[[group_col]], which = "column",
    col = list(group = group_pal)
  )
  ComplexHeatmap::draw(ComplexHeatmap::Heatmap(
    na_mat_subset,
    col = c("0" = "white", "1" = "black"),
    name = "Missing",
    use_raster = FALSE,
    top_annotation = col_ann,
    cluster_rows = TRUE, cluster_columns = TRUE,
    show_row_names = nrow(na_mat_subset) <= 50,
    column_title = "Missing values"
  ))
  grDevices::pdf("plots/missing_values_heatmap.pdf", width = 7, height = 6)
  ComplexHeatmap::draw(ComplexHeatmap::Heatmap(
    na_mat_subset,
    col = c("0" = "white", "1" = "black"),
    name = "Missing", use_raster = FALSE,
    top_annotation = col_ann,
    cluster_rows = TRUE, cluster_columns = TRUE,
    show_row_names = nrow(na_mat_subset) <= 50,
    column_title = "Missing values"
  ))
  invisible(invisible(grDevices::dev.off()))
}
```

### UpSet

```{r}
#| label: fig-vennProteins
#| fig-cap: "UpSet plot: protein overlap across groups."
#| lightbox: true
presence <- lapply(groups, function(g) {
  samples <- col_data$sampleName[col_data[[group_col]] == g]
  rownames(mat_raw)[rowSums(!is.na(mat_raw[, samples, drop = FALSE])) >= 1]
})
names(presence) <- groups
if (length(presence) > 1) {
  UpSetR::upset(UpSetR::fromList(presence), order.by = "freq", nsets = length(groups))
}
```

:::


## Protein Abundance Analysis

::: {.panel-tabset}

### Density

::: {.callout-note collapse="true"}
## About normalization
The density plot displays the protein abundance distribution for all samples.
Left: raw empirical abundances. Right: transformed/normalized abundances.
Major differences between samples may indicate technical biases.
:::

```{r}
#| label: fig-normalized
#| fig-cap: "Protein abundance density. Left: raw. Right: normalized."
#| fig-width: 10
#| fig-height: 5
#| lightbox: true
show_legend <- nrow(col_data) <= 12

p1 <- ggplot2::ggplot(
  long_raw |> dplyr::filter(!is.na(.data$abundance)),
  ggplot2::aes(x = .data$abundance, color = .data$sampleName)
) +
  ggplot2::geom_density() +
  ggplot2::labs(x = "Abundance", y = "Density", title = "Raw") +
  ggplot2::theme_classic() +
  { if (!show_legend) ggplot2::theme(legend.position = "none") }

p2 <- ggplot2::ggplot(
  long_trans |> dplyr::filter(!is.na(.data$abundance)),
  ggplot2::aes(x = .data$abundance, color = .data$sampleName)
) +
  ggplot2::geom_density() +
  ggplot2::labs(x = "Abundance", y = "Density", title = "Normalized") +
  ggplot2::theme_classic() +
  { if (!show_legend) ggplot2::theme(legend.position = "none") }

gridExtra::grid.arrange(p1, p2, nrow = 1)
ggplot2::ggsave("plots/abundance_density.pdf",
  gridExtra::arrangeGrob(p1, p2, nrow = 1), width = 10, height = 5)
```

### CV Analysis

```{r}
#| label: fig-SDViolin
#| fig-cap: "Coefficient of variation (CV) per group."
#| fig-height: 4
#| fig-width: 7
#| lightbox: true
cv_cols <- grep("^CV_", colnames(stats_raw), value = TRUE)
cv_long <- stats_raw |>
  dplyr::select("protein_Id", dplyr::all_of(cv_cols)) |>
  tidyr::pivot_longer(cols = -"protein_Id", names_to = "group", values_to = "CV") |>
  dplyr::mutate(group = gsub("^CV_", "", .data$group)) |>
  dplyr::filter(is.finite(.data$CV))

ggplot2::ggplot(cv_long, ggplot2::aes(x = .data$group, y = .data$CV)) +
  ggplot2::geom_violin(fill = "lightblue", alpha = 0.5) +
  ggplot2::stat_summary(fun = median, geom = "point", size = 2) +
  ggplot2::labs(x = "", y = "CV") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

ggplot2::ggsave("plots/cv_violin.pdf", width = 7, height = 4)
```

```{r}
#| label: tbl-CVtable
medians <- sapply(stats_raw[cv_cols], median, na.rm = TRUE)
cv_tab <- data.frame(group = gsub("^CV_", "", names(medians)),
                     median_CV = round(medians, 4))
rownames(cv_tab) <- NULL
.dt(cv_tab)
```

### Heatmap

::: {.callout-note collapse="true"}
## About the heatmap
Row-scaled protein abundance heatmap using normalized data.
Proteins and samples clustered by euclidean distance.
Grey cells indicate missing values (NAs preserved, not imputed).
:::

```{r}
#| label: fig-heatmap
#| fig-cap: "Protein abundance heatmap (row-scaled, normalized)."
#| fig-width: 7
#| fig-height: 8
#| lightbox: true
mat_hm <- mat_trans
# Need >= 2 non-NA values per row for row-scaling
mat_hm <- mat_hm[rowSums(!is.na(mat_hm)) >= 2, , drop = FALSE]
# Limit to top 500 most variable proteins for readability
if (nrow(mat_hm) > 500) {
  vars <- apply(mat_hm, 1, var, na.rm = TRUE)
  mat_hm <- mat_hm[order(vars, decreasing = TRUE)[1:500], , drop = FALSE]
}
mat_hm_scaled <- t(scale(t(mat_hm)))

# For clustering: after row-scaling, 0 = row mean, so NAs → 0 is the neutral choice
# The heatmap display still shows NAs as grey (na_col)
na_safe_dist <- function(m) { m[is.na(m)] <- 0; dist(m) }

col_ann <- ComplexHeatmap::HeatmapAnnotation(
  group = col_data[[group_col]], which = "column",
  col = list(group = group_pal)
)

hm <- ComplexHeatmap::Heatmap(
  mat_hm_scaled,
  name = "z-score",
  na_col = "grey80",
  use_raster = FALSE,
  top_annotation = col_ann,
  show_row_names = FALSE,
  show_column_names = TRUE,
  clustering_distance_rows = na_safe_dist,
  clustering_distance_columns = na_safe_dist,
  cluster_rows = TRUE, cluster_columns = TRUE,
  column_title = "Protein abundances (row-scaled)"
)
ComplexHeatmap::draw(hm)

grDevices::pdf("plots/abundance_heatmap.pdf", width = 7, height = 8)
ComplexHeatmap::draw(hm)
invisible(grDevices::dev.off())
```

### PCA

```{r}
#| label: fig-pca
#| fig-cap: "PCA of normalized protein abundances (PC1 vs PC2)."
#| fig-width: 7
#| fig-height: 7
#| lightbox: true
mat_pca <- mat_trans[complete.cases(mat_trans), , drop = FALSE]
pca_res <- prcomp(t(mat_pca), center = TRUE, scale. = TRUE)
var_expl <- summary(pca_res)$importance[2, 1:2] * 100
pca_df <- data.frame(
  PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2],
  sampleName = rownames(pca_res$x)
)
pca_df <- dplyr::left_join(pca_df, col_data, by = "sampleName")

ggplot2::ggplot(pca_df, ggplot2::aes(
  x = .data$PC1, y = .data$PC2,
  color = .data[[group_col]], label = .data$sampleName
)) +
  ggplot2::geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3) +
  ggplot2::scale_color_manual(values = group_pal) +
  ggplot2::labs(
    x = sprintf("PC1 (%.1f%%)", var_expl[1]),
    y = sprintf("PC2 (%.1f%%)", var_expl[2])
  ) +
  ggplot2::theme_classic()

ggplot2::ggsave("plots/pca.pdf", width = 7, height = 7)
```

:::


## Differential Expression

::: {.panel-tabset}

### Volcano

::: {.callout-note collapse="true"}
## About the volcano plot
Volcano plot showing $-\log_{10}$(FDR) vs log2 fold-change.
Red dashed line: FDR threshold. Green dashed lines: difference threshold.
Orange dots: estimates based on missing value imputation.
:::

```{r}
#| label: fig-volcanoplot
#| fig-cap: "Volcano plot per contrast."
#| fig-width: 9
#| fig-height: 7
#| lightbox: true
ctr_all <- ctr_all |>
  dplyr::mutate(
    neg_log10_fdr = -log10(.data$FDR),
    color = ifelse(grepl("Imputed", .data$modelName, ignore.case = TRUE),
                   "Imputed", "Modelled")
  )

ggplot2::ggplot(ctr_all, ggplot2::aes(
  x = .data$diff, y = .data$neg_log10_fdr, color = .data$color
)) +
  ggplot2::geom_point(alpha = 0.5, size = 1) +
  ggplot2::scale_color_manual(values = c("Modelled" = "black", "Imputed" = "orange")) +
  ggplot2::geom_hline(yintercept = -log10(params$fdr_threshold),
                      linetype = "dashed", color = "red") +
  ggplot2::geom_vline(xintercept = c(-params$diff_threshold, params$diff_threshold),
                      linetype = "dashed", color = "green") +
  ggplot2::facet_wrap(~contrast) +
  ggplot2::labs(x = "diff (log2 FC)", y = "-log10(FDR)", color = "Model") +
  ggplot2::theme_classic()

ggplot2::ggsave("plots/volcano.pdf", width = 9, height = 7)
```

### Results Table

::: {.callout-note}
## Analysis Settings

- **Formula**: `r meta$formula$formula[1]`
- **FDR threshold**: `r params$fdr_threshold`
- **Difference threshold**: `r params$diff_threshold`
:::

```{r}
#| label: tbl-contrasts
.dt(meta$contrasts)
```

```{r}
#| label: tbl-tableAllProt
bb <- ctr_all |>
  dplyr::select(dplyr::any_of(c(
    "protein_Id", "description", "gene_name",
    "contrast", "modelName", "diff", "FDR"
  )))
.dt(bb) |> DT::formatRound(columns = c("FDR", "diff"), digits = 3)
```

### Significant Proteins

```{r}
#| label: tbl-nrsignificant
mx <- ctr_all |>
  dplyr::mutate(
    passes = abs(.data$diff) > params$diff_threshold &
      .data$FDR < params$fdr_threshold
  ) |>
  dplyr::group_by(.data$contrast) |>
  dplyr::summarise(
    n = dplyr::n(),
    Significant = sum(.data$passes, na.rm = TRUE),
    `Not Significant` = dplyr::n() - sum(.data$passes, na.rm = TRUE),
    .groups = "drop"
  )
.dt(mx)
```

```{r}
#| eval: !expr nrow(ctr_signif) > 0
#| label: tbl-SigPrey
sig_display <- ctr_signif |>
  dplyr::select(dplyr::any_of(c(
    "protein_Id", "description", "gene_name",
    "contrast", "modelName", "diff", "FDR"
  )))
.dt(sig_display) |> DT::formatRound(columns = c("FDR", "diff"), digits = 3)
```

### Significant Heatmap

```{r}
#| eval: !expr nrow(ctr_signif) > 0
#| label: fig-sigHeatmap
#| fig-cap: "Heatmap of significant proteins (row-scaled, normalized). Grey = missing."
#| fig-width: 7
#| fig-height: 8
#| lightbox: true
sig_ids <- unique(ctr_signif$protein_Id)
mat_sig <- mat_trans[sig_ids, , drop = FALSE]
mat_sig <- mat_sig[rowSums(!is.na(mat_sig)) >= 2, , drop = FALSE]

if (nrow(mat_sig) > 1) {
  mat_sig_scaled <- t(scale(t(mat_sig)))
  col_ann <- ComplexHeatmap::HeatmapAnnotation(
    group = col_data[[group_col]], which = "column",
    col = list(group = group_pal)
  )
  hm_sig <- ComplexHeatmap::Heatmap(
    mat_sig_scaled,
    name = "z-score", na_col = "grey80", use_raster = FALSE,
    top_annotation = col_ann,
    show_row_names = nrow(mat_sig_scaled) <= 30,
    clustering_distance_rows = na_safe_dist,
    clustering_distance_columns = na_safe_dist,
    cluster_rows = TRUE, cluster_columns = TRUE,
    column_title = "Significant proteins (row-scaled)"
  )
  ComplexHeatmap::draw(hm_sig)
  grDevices::pdf("plots/significant_heatmap.pdf", width = 7, height = 8)
  ComplexHeatmap::draw(hm_sig)
  invisible(invisible(grDevices::dev.off()))
}
```

### UpSet

```{r}
#| eval: !expr nrow(ctr_signif) > 0 && length(unique(ctr_signif$contrast)) > 1
#| label: fig-vennSig
#| fig-cap: "UpSet plot: significant proteins across contrasts."
#| lightbox: true
xx <- split(ctr_signif$protein_Id, ctr_signif$contrast)
if (length(xx) > 1) {
  UpSetR::upset(UpSetR::fromList(xx), order.by = "freq")
}
```

:::


## Additional Analysis

::: {.callout-note}
## Output Files
The zip file (if provided) contains:

- **Excel file** with sheets: annotation, raw/normalized abundances, differential expression results, missing information, protein variances
- **ORA files** (`ORA_background.txt`, `ORA_<contrast>.txt`) for over-representation analysis
- **GSEA rank files** (`GSEA_<contrast>.rnk`) for gene set enrichment analysis
:::

Recommended tools for functional enrichment:

- [STRING-DB](https://string-db.org/) — protein-protein interaction networks
- [DAVID](https://david.ncifcrf.gov/home.jsp) — functional annotation
- [WebGestalt](http://www.webgestalt.org) — over-representation and GSEA


## Session Info

```{r}
pander::pander(sessionInfo())
```

:::

---

*Generated from `Grp2Analysis_V2_SE.qmd` in the `prolfquapp` R package.*
