---
title: "Differential Expression Analysis."
author: "Functional Genomics Center Zurich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
  bookdown::html_document2:
    toc: true
  pdf_document: 
    toc: true
header-includes: 
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[CO,CE]{Group Comparison}
  \fancyfoot[CO,CE]{\textbf{FGCZ} - www.fgcz.ch - 2018}
  \fancyfoot[LE,RO]{\thepage} 
params:
  grp: NULL
vignette: >
  %\VignetteIndexEntry{FGCZ Two-Group Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)



grp2 <- params$grp
```


# B-fabric related information

This report is stored in the LIMS system [_bfabric_](https://fgcz-bfabric.uzh.ch) [@panse2022bridging] in
__project__: `r grp2$project_spec$project_Id`, __order__: `r grp2$project_spec$order_Id`, `r if(!is.null(grp2$project_spec$workunit_Id)){ "with the workunit name :" }`  `r grp2$project_spec$workunit_Id`.


The metabolite identification and quantification were performed using: 
`r if(!is.null(grp2$software)){grp2$software} else {"<Specify parameter grp2$software>"}`. The input file can be downloaded from here : [`r grp2$project_spec$input_URL`](`r grp2$project_spec$input_URL`). The set of parameters used to run the quantification software can be retrieved from.


# Introduction

The differential expression analysis verifies if the __difference__ between normalized empirical metabolite abundances measured in two groups is significantly non-zero.
To make the test as sensitive and specific these empirical abundances $A$ are further $log_2$ transformed and scaled to make them compatible with the statistical test procedure [@valikangas2018systematic]. Therefore, we obtain a scale free $\log_2$ transformed normalized protein abundances $\log_2(A)$ for a sample.

For unpaired experiments the difference $\Delta$ between group $a$ and $b$, for a specific protein is estimated by:  

$$
\Delta = \frac{1}{n}\sum_{i=1}^n \log_2(A^a_i) - \frac{1}{m}\sum_{j=1}^{m}\log_2(A^b_j)
$$


where $A^a_i$ - is the normalized metabolite abundance of sample $i$ in the group $a$ of $n$ samples, while $A^b_j$ is the metabolite abundance of sample $j$ in group $b$ of $m$ samples.

For paired experiment the difference is estimated by:


$$
\Delta = \frac{1}{n}\sum_{i=1}^n \log_2(A^a_i) - \log_2(A^b_{i})
$$


where $n$ is the number of subjects, each treated with $a$ and $b$.

Of note, when comparing two samples $a$ and $b$ the difference of logarithms equals to the logarithm of the ratio (exponent rule):

$$
\log(A^a) - \log(A^b) = \log(\frac{A^a}{A^b})
$$


, it is called $\log$-ratio or $\log$ fold-change (logFC).

The estimated differences $\Delta$ have an associated error $\epsilon$. Therefore, the differential expression analysis must test if the difference is significantly nonzero. 

We run a set of functions implemented in the R package _[prolfqua]_ [@prolfquawolski2023] to normalize the data, generate visualizations, and to compute differential expression analysis. To future improve the power of the differential expression test the protein variances are moderated [@Smyth2004], i.e. the individual protein variances are updated using a variance prior estimated from all the proteins in the experiment.

# Results

Table \@ref(tab:samples) shows the number of samples assigned to each group while Table \@ref(tab:annotation) shows the names of the files assigned to the group.

```{r samples, eval=TRUE}

tab <- data.frame(table(grp2$RES$lfqData$factors()[grp2$RES$lfqData$config$table$factor_keys_depth()]))
colnames(tab)[length(colnames(tab))] <- "# samples"
knitr::kable(tab, caption = "Nr of samples assigned to each group.")

```


```{r annotation, eval=TRUE}
ld <- grp2$RES$lfqData
fx <- ld$factors()
sr <- ld$get_Summariser()
xx <- sr$hierarchy_counts_sample(nr_children = 1)
xx$isotopeLabel <- NULL
xx <- xx |> dplyr::rename(nr_1 = ld$config$table$hierarchy_keys())
x2 <- sr$hierarchy_counts_sample(nr_children = 2)
x2$isotopeLabel <- NULL
x2 <- x2 |> dplyr::rename(nr_2 = ld$config$table$hierarchy_keys())

fx <- dplyr::left_join(fx,xx, by = ld$config$table$sampleName)
fx <- dplyr::left_join(fx,x2, by = ld$config$table$sampleName)


knitr::kable(fx, caption = "LC-MS samples annotation table. The content of the sampleName column is used as a short form to plot labels. The group to which a sample is assigned to is shown in the column group.")

```

## Metabolite quantification and identification

Figure \@ref(fig:nrPerSample) shows the number of quantified metabolites.

```{r nrPerSample, fig.cap="Number of identified proteins across samples.", fig.with=10, fig.height=7}
sum <- grp2$RES$lfqData$get_Summariser()
sum$plot_hierarchy_counts_sample(nr_children = 1)

```

## Missing Value Analysis

The absence of a metabolite measurement in a sample might be biologically relevant or might point to technical problems. Significant differences in the set of metabolites observed in the samples within a group typically indicate either technical problems or excessive biological variability. If for instance, one sample out of ten has a different set of proteins, it is likely an outlier and can be removed from the analysis. [TODO: Unclear]If the differences between the groups are significant but within the groups are small, this might systematically bias the difference estimates, i.e., produce false-positive or false-negative test results.


```{r countProtWithNAs}
res <- grp2$RES$lfqData$to_wide(as.matrix = TRUE)$data
res[!is.na(res)] <- 0
res[is.na(res)] <- 1
allrows <- nrow(res)
res <- res[apply(res,1, sum) > 0, , drop = FALSE]
allNas <- nrow(res)
```


```{r prepHeatmap}
pl <- grp2$RES$lfqData$get_Plotter()
nah <- pl$NA_heatmap()
```

A dichotomous view of the data can be constructed by transforming protein abundance estimates into present/absent calls (Figure \@ref(fig:naHeat) ). The heatmap shows only proteins with at least one missing value. There are `r allNas` proteins with at least one missing value in the data, which is (`r allNas/allrows * 100` %).

We expect that samples in the same group are more similar and cluster together, i.e., they are in the same branch of the dendrogram.

(ref:naHeat) Protein abundance heatmap (rows indicate proteins, columns indicate samples) showing missing protein abundance estimates across data set. Rows and columns are grouped based on the Minkowski distance using hierarchical clustering. White: Metabolite is observed, black: Metabolite is not observed.

```{r naHeat, fig.width=7, fig.height=6, dpi=300, fig.cap="(ref:naHeat)", eval=TRUE}
nah
```

Using Figure \@ref(fig:vennProteins) we examine if we see the same metabolite in each group. We say a metabolite is unobserved in the group if it is absent in all samples and is present otherwise. A significant overlap among groups allows more precisely estimating the metabolite abundance differences between the groups.

(ref:vennProteins) Venn diagram showing the number of proteins present in each group and in all possible intersections among groups. 

```{r vennProteins, fig.cap="(ref:vennProteins)"}
pups <- prolfqua::UpSet_interaction_missing_stats(grp2$RES$lfqData$data, grp2$RES$lfqData$config, tr = 1)
UpSetR::upset(pups$data, order.by = "freq", nsets = pups$nsets)

```


## Protein Abundance Analysis

The density plot (Figure \@ref(fig:normalized) left panel)  displays the metabolite abundance distribution for all data set samples. Mayor differences between samples could hint that the individual metabolite abundance values are affected by technical biases. These biases might need to be corrected to separate them from biological effects. The right panel of Figure \@ref(fig:normalized) shows the distribution of the transformed and scaled normalized empirical metabolite abundances. Normalization is applied to remove systematic differences in metabolite abundances due to different sample concentrations or amounts of sample loaded on a column. However, in the presence of a large proportion of missing data, normalization potentially amplifies systematic errors.


```{r normalization, results = 'asis'}
if (grp2$processing_options$transform == "robscale") {
  cat("To do this the z-score of the $\\log_2$ transformed metabolite abundances are computed.",
      "Because we need to estimate the metabolite differences on the original scale, we have to multiply the $z$-score by the average standard deviation of all the $N$ samples in the experiment.",
  "After normalization all samples have an equal mean and variance and a similar distribution.")
} else if (grp2$processing_options$transform == "vsn") {
  cat("To do this the variance stabilizing normalization (vsn) was applied [@HuberVSN2002].")
} else if (grp2$processing_options$transform == "none") {
  cat("However, in some circumstances it is advisable not to normalize the data, e.g. in case of affinity purification experiments, or when the requirements of sufficient similarity among samples are not met.")
}

```

(ref:normalized) Kernel density function showing the distribution of metabolite abundances in all samples. Left panel : Empirical metabolite abundance of all samples in the dataset. Right panel: Normalized empirical metabolite abundances of all samples in the dataset.

```{r normalized,  fig.width=8, fig.height=5,dpi=300, fig.cap="(ref:normalized)", eval=TRUE}

pl <- grp2$RES$lfqData$get_Plotter()

showlegend <- if (nrow(grp2$RES$lfqData$factors()) > 12) {FALSE} else {TRUE}
p1 <- pl$intensity_distribution_density(legend = showlegend)

plTransformed <- grp2$RES$transformedlfqData$get_Plotter()
p2 <- plTransformed$intensity_distribution_density(legend = showlegend) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))

gridExtra::grid.arrange(p1, p2, nrow = 1)

```

The median coefficient of variation (CV) of a group of samples or all samples in the dataset, can be used to compare the experiment with other experiments [@piehowski2013sources]. For example, the median CV for high-performance liquid chromatography experiments ranges from 2% to 35% depending on the biological samples studied, the chromatography method used, label-free or labelled quantification [@taverna2021critical].


Figure \@ref(fig:SDViolin), show the coefficients of variations (CV) for all metabolites computed on not normalized data. Ideally the within group CV should be smaller than the CV of all samples. 


(ref:SDViolin) Distribution of coefficient of variation (CV) within each groups and in the entire experiment (all).

```{r SDViolin, fig.height=3, fig.width=7, fig.cap="(ref:SDViolin)", eval=TRUE}
stR <- grp2$RES$lfqData$get_Stats()
pA <- stR$violin() + ggplot2::theme_classic() + ggplot2::labs(tag = "A")
pA

```

The Table \@ref(tab:CVtable) shows the median CV of all groups and across all samples (all).

```{r CVtable, eval=TRUE}
resR <- stR$stats_quantiles()
C <- dplyr::bind_rows(CV = resR$wide  |> dplyr::filter(probs == 0.5) |> round(digits = 2))
C <- C |> dplyr::mutate( what = c("CV"), .before = 1 )
C$probs <- NULL
knitr::kable(C, caption = 'Median of coefficient of variation (CV).')

```



```{r generateHeatmaps, inlcude = FALSE}
pl <- grp2$RES$transformedlfqData$get_Plotter()
ph <- pl$heatmap()

```

The metabolite abundance heatmap (Figure \@ref(fig:heatmap)) groups the metabolite and samples using unsupervised hierarchical clustering. Distances between metabolites and samples are computed using normalized metabolite abundances. Metabolites with a large proportion of missing observations are not shown in this heatmap, because for these metabolites no distances can be computed. Metabolites and samples showing similar abundances are grouped and shown in adjacent rows and columns respectively.


(ref:heatmap) Protein abundance heatmap (rows indicate proteins, columns indicate samples) showing the row scaled $\log_2$ transformed protein abundance value. Co-clustering (hierarchical complete linkage, euclidean distance) of samples and proteins was used.



```{r heatmap, fig.width=7, fig.height=8, dpi=300, fig.cap="(ref:heatmap)", eval=TRUE}
ph
```

We use principal component analysis (PCA) to transform the high dimensional space defined by all metabolites into a two-dimensional one, containing most of the information. Plot \@ref(fig:pca) shows the location of the samples according to the first and second principal component, which explain most of the variance in the data. Samples close in the PCA plot are more similar than those farther apart.

(ref:pca)  Plot of first and second principal component (PC1 and PC2) of principal component analysis (PCA). Normalized abundances were used as input.

```{r pca, fig.cap = "(ref:pca)" , fig.width=7, fig.height=7 }
grp2$RES$transformedlfqData$get_Plotter()$pca_plotly()
```

# Differential Expression Analysis

The method used to test for differential expression consists of several steps: First a linear model that explains the observed protein abundances using the grouping of the samples is fitted using the R function _lm_ to each protein:

`r grp2$RES$formula`.

Secondly, the difference between the group is computed (Table \@ref(tab:contrtable)).

```{r contrtable}
df <- data.frame(name = names(grp2$pop$Contrasts), contrast = grp2$pop$Contrasts)
rownames(df) <- NULL
knitr::kable(df, caption = "Name of difference (Contrast), and formula used to compute it.")

```


and a null hypothesis significance test (NHST) is conducted, where the null hypothesis is that the protein is not differentially expressed [@faraway2004linear]. 

If there are no abundances measured in one of the groups for some metabolites, we assume the observations are missing because the metabolite abundance is below the detection limit. Furthermore, to make it explicit for which metabolites we did impute the unobserved group mean, we label them with `Imputed_Mean` (see table in Figures \@ref(fig:tableAllProt) column `modelName`) and visualize them with gray dots in Figure \@ref(fig:volcanoplot). Finally, those metabolites with a sufficiently large number of observations are labeled with `Linear_Model_Moderated`.

Next, to increase the power of the analysis variance shrinkage is performed [@Smyth2004]. Finally, the false discovery rate (FDR) using the Benjamini-Hochberg procedure is computed [@benjamini1995controlling].
The FDR is the expected proportion of false discoveries in a list of metabolites, and can be used to select candidates for follow up experiments. FDR thresholds commonly used are 5, 10 or 25%. By filtering the metabolites using a FDR threshold of `r grp2$processing_options$FDR_threshold*100` % we can expect this proportion of false positives in the list and `r (1 - grp2$processing_options$FDR_threshold) * 100` % truly differentially expressed metabolites. Because we do not know which of them are true positives follow up experiments are necessary.

The table (Figure \@ref(fig:tableAllProt)) summarizes the differential expression analysis results by providing the following information:

```{r}
hkey <- grp2$RES$lfqData$config$table$hierarchy_keys()
```

- `r if(hkey == "metabolite_Id"){ paste0(hkey, " - unique metabolite identifier")} else {hkey}`
- description - information about the metabolite
- contrast - name of the comparison
- modelName - name of the method to estimate differences : Imputed_mean or Linear_Model_Moderated
- FDR - false discovery rate
- diff - difference between groups.

The volcano plot \@ref(fig:volcanoplot) helps to identify metabolites with a large differences among groups and a low FDR. The significance dimension is a $-\log_{10}$ transformed FDR, i.e., small values of FDR become large after transformation. Promising candidate metabolites are found in the upper right and left sector of the plot.


```{r setupVolcano, echo = FALSE}
datax <- as.data.frame(grp2$RES$contrastsData)
```


```{r tableAllProt, fig.cap="Differential expression analysis results of all metabolites."}
bb <- datax |> dplyr::select(
  all_of(c(hkey,
           "description",
           "contrast",
           "modelName",
           "diff",
           "FDR" )))
bb <- crosstalk::SharedData$new(bb,  as.formula(paste0("~",hkey)), group = "BB")

DT::datatable(bb, filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE) ) |>
  DT::formatRound(columns = c('FDR', 'diff'), digits = 3)

```

(ref:volcanoplot)  Volcano plot showing $-\log_{10}$ transformed FDR as function of the difference between groups. The red line indicates the $-log_{10}(FDR)$ of FDR = `r grp2$processing_options$FDR_threshold`, while the green lines represent the the difference of minus and plus `r grp2$processing_options$diff_threshold`. With orange dots differences and FDRs estimated using missing value imputation are shown.


```{r volcanoplot, fig.cap = "(ref:volcanoplot)", fig.width=9, fig.height=7, include = TRUE, eval = TRUE}

palette <- c("black","orange")
palette <- setNames(palette, c("Linear_Model_moderated", "Imputed_Mean_moderated"))  
xd <- prolfqua::volcano_plotly( 
  datax ,
  proteinID = hkey,
  effect = "diff",
  significance = "FDR",
  contrast = "contrast",
  color = "modelName",
  palette = palette,
  xintercept =  c(-grp2$processing_options$diff_threshold,grp2$processing_options$diff_threshold),
  yintercept = grp2$processing_options$FDR_threshold,
  title_size = 10,
  group = "BB")

xd <- lapply(xd, plotly::highlight , off = "plotly_doubleclick")
nrow <- ceiling(length(xd) / 4)
plotly::subplot(xd, shareX = TRUE, shareY = TRUE, nrows = nrow)

```

## Differentially Expressed Metabolites

Here we use the FDR threshold of `r grp2$processing_options$FDR_threshold` and a difference threshold of `r grp2$processing_options$diff_threshold` to select differentially expressed metabolites. Table \@ref(tab:nrsignificant) summarizes the number of significant calls.

```{r nrsignificant, results="markup", eval=TRUE}
library(dplyr)
mx <- datax |> dplyr::mutate(passes = abs(diff) >  grp2$processing_options$diff_threshold  & FDR < grp2$processing_options$FDR_threshold)
x <- mx |> dplyr::group_by(contrast) |> 
  summarize(n = n(), Significant = sum(passes, na.rm = TRUE), "Not Significant" = n() - sum(passes, na.rm = TRUE))

mycap <- paste0("Number of not significant and significant metabolites."  )
knitr::kable(x, caption = mycap)

```



```{r areThereSig}
datax_signif <- grp2$RES$contrastsData_signif
showSignificant <- TRUE
if (nrow(datax_signif) == 0) {
  showSignificant <- FALSE
}

```


```{r results = 'asis', eval = showSignificant}
cat("The table shown in Figure \\@ref(fig:SigPrey) lists all the significant metabolites.")
```

(ref:SigPrey) Significant metabolites obtained by applying the difference and FDR thresholds.


```{r SigPrey, fig.cap= "(ref:SigPrey)", eval = showSignificant}
ctdata <- datax_signif |> dplyr::select(all_of(c(hkey ,  "description", "contrast", "modelName", "diff", "FDR"  ))) |> as.data.frame()

sig <- crosstalk::SharedData$new(ctdata,  as.formula( paste0("~",hkey)), group = "BB")
DT::datatable(sig, filter = "bottom",
  extensions = "Scroller",
  style = "auto",
  class = "compact",
  options = list(deferRender = TRUE,
                 scrollY = 300,
                 scrollX = 400,
                 scroller = TRUE) ) |> 
  DT::formatRound(columns = c('FDR', 'diff'), digits = 3)
```




```{r prepareHeatmap, eval = showSignificant}
signif <- grp2$RES$transformedlfqData$get_copy()
signif <- signif$get_subset(datax_signif)
if (signif$hierarchy_counts()[2] > 30) {
  rownames = FALSE
} else {
  rownames = TRUE
}
sigheat <- signif$get_Plotter()$raster(rownames = rownames)

```


```{r makeText, eval = showSignificant, results = 'asis'}

cat(paste("Furthermore, Figure \\@ref(fig:sigroteins) shows a heatmap of log2 transformed metabolite abundances of all significant calls."))

```

(ref:sigroteins) Heatmap showing the $\log_2$ transformed metabolite abundances for metabolites which pass the FDR and difference thresholds.

```{r sigroteins, fig.cap="(ref:sigroteins)", eval = showSignificant}
sigheat
```

(ref:vennDiagramSig) Venn diagram showing the number of significant metabolites for each contrast and their intersections.


```{r vennDiagramSig, fig.cap="(ref:vennDiagramSig)", eval = showSignificant}
xx <- split(datax_signif[[hkey]], datax_signif$contrast)
if (length(xx) > 1) {
  UpSetR::upset(UpSetR::fromList(xx))
}

```


# Additional Analysis

The zip file contains an excel file __DE_Groups_vs_Controls.xlsx__. All the figures can be recreated using the data in the excel file. The Excel file contains the following spreadsheets:

- __annotation__ - the annotation of the samples in the experiment
- __raw_abundances__ table with empirical metabolite abundances.
- __normalized_abundances__ table with normalized metabolite abundances. 
- __raw_abundances_matrix__ A table where each column represents a sample and each row represents a metabolite and the cells store the empirical metabolite abundances.
- __normalized_abundances_matrix__ A table where each column represents a sample and each row represents a metabolite and the cells store the empirical metabolite abundances.
- __diff_exp_analysis__ A table with the results of the differential expression analysis. For each metabolite there is an row containing the estimated difference between the groups, the false discovery rate FDR, the 95% confidence interval, the posterior degrees of freedom. 
- __missing_information__ - spreadsheet containing information if a metabolite is present (1) or absent in a group (0).
- __metabolite_variances__ - spreadsheet which for each metabolite shows the variance (var) or standard deviation (sd) within a group, the number of samples (n) and the number of observations (not_na) as well as the group average intensity (mean).


# Session Information

```{r sessionInfo}
pander::pander(sessionInfo())
```

# References {-}

<div id="refs"></div>


# Glossary

- groups - different treatments, genotypes etc.
- diff (difference) - it is the difference of the protein abundance estimate of two groups
- FDR - false discovery rate

